---
title: "Reproducible flower-colour data preparation"
output: html_document
params:
  input: "Data_S1.csv"
  sdm_dir: "sdm"
  accepted_qc_status: "ok,manual_review_required"
  colour_method: "primary"
  grain: "photo"
  site_id: !r NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)

find_root <- function(start = getwd()) {
  path <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    if (file.exists(file.path(path, "Data_S1.csv")) &&
        file.exists(file.path(path, "R", "analysis_core.R"))) {
      return(path)
    }
    parent <- dirname(path)
    if (identical(parent, path)) stop("Project root was not found.")
    path <- parent
  }
}

project_root <- find_root()
source(file.path(project_root, "scripts", "run_analysis.R"), local = environment())

resolve_report_path <- function(path) {
  path <- path.expand(as.character(path)[1L])
  if (grepl("^(/|[A-Za-z]:[/\\\\])", path)) path else file.path(project_root, path)
}
```

## Scope

This report validates and prepares the repository's standard input. The unit of
analysis is a photograph by default. Exact duplicate coordinates are reported,
but never silently collapsed. Site-level analysis requires an explicit site ID.
It is a lightweight preparation report; the canonical multi-method analysis
with the complete public raster stack, common cohorts, blocked validation, and
run manifests is `scripts/run_reanalysis.R`.

The colour response below is an image-colour axis, not a direct measurement of
pigment concentration. `Bombus_suitability_sum` is the sum of five continuous
SDM predictions, not observed species richness.

## Input contract and grain

```{r input}
input_path <- resolve_report_path(params$input)
report_output_dir <- tempfile("code-s3-")
dir.create(report_output_dir)
analysis_args <- c(
  "--input", input_path,
  "--sdm-dir", resolve_report_path(params$sdm_dir),
  "--accepted-qc-status", params$accepted_qc_status,
  "--colour-method", params$colour_method,
  "--grain", params$grain,
  "--output", file.path(report_output_dir, "prepared.csv"),
  "--loadings-output", file.path(report_output_dir, "loadings.csv")
)
if (identical(params$grain, "site")) {
  if (is.null(params$site_id) || !nzchar(params$site_id)) {
    stop("The site grain requires params$site_id.")
  }
  analysis_args <- c(analysis_args, "--site-id", params$site_id)
}
analysis_result <- run_analysis_cli(
  analysis_args,
  project_root = project_root
)
analysis_data <- analysis_result$data
colour_pca <- analysis_result$colour_pca
list(
  input_grain_audit_before_qc = analysis_result$input_grain_audit,
  colour_qc_audit = analysis_result$qc_audit,
  photo_coordinate_identity_audit = analysis_result$photo_coordinate_audit,
  colour_method_audit = analysis_result$colour_method_audit,
  analysis_grain_audit_after_qc = analysis_result$grain_audit
)
```

When `qc_status` is present, only the configured statuses enter colour PCA or
raster extraction. The preparation default retains both `ok` and
`manual_review_required`: several automated flags depend on colour itself, so
silently dropping all flagged photographs could create response-dependent
selection. This default is not manual approval. Inference should use completed
visual review or report both inclusive and strict `accepted_qc_status: "ok"`
sensitivity analyses. Excluded statuses and counts are displayed above.

An unresolved exact-photo duplicate assigned to multiple coordinates is a
different problem from colour-dependent QC. Both conflicting records are
always excluded from analysis preparation until the source identity is
resolved; they remain in `Data_S1.csv` for auditability.

The `primary` colour method uses the published median `R/G/B` columns. Optional
sensitivity methods copy one explicitly named candidate triplet into `R/G/B`;
the original values remain in `primary_R/G/B`. Records missing any selected
candidate channel are excluded and counted, never filled from another method.

## sRGB to CIELAB and sign-fixed colour PCA

RGB values are required to use the 0--255 sRGB scale. Missing colours remain in
the table as missing values; partially missing RGB triples are rejected.

```{r colour}
data.frame(
  variable = names(colour_pca$loadings),
  PC1_loading = unname(colour_pca$loadings),
  explained_variance_PC1 = colour_pca$explained_variance,
  anchor = colour_pca$anchor,
  row.names = NULL
)
```

The requested grain is established before fitting PCA. Repeated photographs
therefore cannot silently reweight a site-grain PCA. The PC1 sign is fixed so
that its loading on a* is positive, preventing an arbitrary sign reversal.

## Five-species Bombus suitability

```{r bombus}
species <- bombus_species_columns()
sdm_validation <- analysis_result$sdm_validation
if (!isTRUE(sdm_validation$model_reproducible)) {
  message(
    "SDM structure is valid, but model-generation provenance is ",
    "legacy_unverifiable."
  )
}

data.frame(
  records = nrow(analysis_data),
  complete_five_species = sum(analysis_data$Bombus_predictions_n == 5L),
  incomplete_predictions = sum(analysis_data$Bombus_predictions_n < 5L),
  finite_suitability_sum = sum(is.finite(analysis_data$Bombus_suitability_sum))
)
```

An all-missing SDM row remains `NA`; it is never converted to a false zero.
Raster results are reattached by `record_id`, so filtering coordinates cannot
shift values onto another photograph.

## Explicit analysis grain

```{r grain}
data.frame(grain = params$grain, analysis_rows = nrow(analysis_data))
```

## Heavy spatial models

INLA and qGAM are deliberately not run by this report. A replacement model must
first specify a Japan-appropriate metric projection, justified mesh sensitivity,
PC priors, consistent standardization, spatially blocked validation, and a joint
treatment of population density and spatial effects. If a variance partition is
reported, it must use posterior draws and include the
`2 * Cov(fixed, spatial)` contribution via
`variance_decomposition_with_covariance()`.

```{r heavy-model-status}
message("INLA/qGAM status: intentionally disabled pending a reviewed joint model specification.")
```
