---
title: "Campanula punctata: final analysis"
output:
  html_document:
    toc: false
    code_folding: hide
    self_contained: false
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
```

```{r packages}
if (!requireNamespace("patchwork", quietly = TRUE)) {
  utils::install.packages("patchwork", repos = "https://cloud.r-project.org")
}

suppressPackageStartupMessages({
  library(patchwork)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(sf)
  library(terra)
  library(lavaan)
  library(tibble)
  library(INLA)
  library(qgam)
  library(mgcv)
  library(ggplot2)
  library(forcats)
  library(purrr)
  library(broom)
  library(jsonlite)
  library(digest)
})

set.seed(42)
INLA::inla.setOption(num.threads = "1:1")

if (requireNamespace("RhpcBLASctl", quietly = TRUE)) {
  RhpcBLASctl::blas_set_num_threads(1)
  RhpcBLASctl::omp_set_num_threads(1)
}

RNGkind(kind = "Mersenne-Twister", normal.kind = "Inversion")

z <- function(x) as.numeric(scale(x))


out_root <- path.expand("~/zui_results")
dir.create(out_root, recursive = TRUE, showWarnings = FALSE)

cmp_dir      <- file.path(out_root, "diagnostics")
cmp_plotdir  <- file.path(cmp_dir, "plots")
top_dir      <- file.path(out_root, "supplement_sensitivity")
best_dir     <- file.path(out_root, "natural_baseline_final")
best_plotdir <- file.path(best_dir, "plots")
gam_dir      <- file.path(out_root, "qGAM_best_model_final")
pub_dir      <- file.path(out_root, "publication_materials")
fig_dir      <- file.path(pub_dir, "figures")
table_dir    <- file.path(pub_dir, "tables")
text_dir     <- file.path(pub_dir, "manuscript_text")
repro_dir    <- file.path(pub_dir, "reproducibility")

dir.create(cmp_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(cmp_plotdir, recursive = TRUE, showWarnings = FALSE)
dir.create(top_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(best_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(best_plotdir, recursive = TRUE, showWarnings = FALSE)
dir.create(gam_dir, recursive = TRUE, showWarnings = FALSE)
for (d in c(pub_dir, fig_dir, table_dir, text_dir, repro_dir)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

save_pub_plot <- function(plot, filename, width, height) {
  print(plot)
  ggplot2::ggsave(file.path(fig_dir, paste0(filename, ".pdf")), plot,
                  width = width, height = height, device = cairo_pdf)
  ggplot2::ggsave(file.path(fig_dir, paste0(filename, ".tiff")), plot,
                  width = width, height = height, dpi = 600,
                  compression = "lzw")
  invisible(plot)
}

capture_warnings <- function(expr) {
  messages <- character()
  value <- withCallingHandlers(
    expr,
    warning = function(w) {
      messages <<- c(messages, conditionMessage(w))
      invokeRestart("muffleWarning")
    }
  )
  list(value = value, warnings = unique(messages))
}
```

## Colour

```{r load-color}
in_csv <- "~/Desktop/Data_S1.csv"
df <- read.csv(in_csv, stringsAsFactors = FALSE, check.names = FALSE)

stopifnot(all(c("longitude", "latitude") %in% names(df)))

if (!("sample_id" %in% names(df))) {
  df$sample_id <- if ("observation_id" %in% names(df)) df$observation_id else seq_len(nrow(df))
}

rgb_cols <- if (all(c("median_R", "median_G", "median_B") %in% names(df))) {
  c("median_R", "median_G", "median_B")
} else {
  stopifnot(all(c("R", "G", "B") %in% names(df)))
  c("R", "G", "B")
}

rgb_primary <- as.data.frame(lapply(df[, rgb_cols, drop = FALSE], as.numeric))
names(rgb_primary) <- c("R_primary", "G_primary", "B_primary")
stopifnot(all(vapply(rgb_primary, function(x) all(is.na(x) | (x >= 0 & x <= 255)), logical(1))))
df <- cbind(df, rgb_primary)

srgb_to_lab_d65 <- function(r, g, b) {
  rgb <- pmin(pmax(cbind(as.numeric(r), as.numeric(g), as.numeric(b)) / 255, 0), 1)
  lin <- ifelse(rgb <= 0.04045, rgb / 12.92, ((rgb + 0.055) / 1.055)^2.4)
  xyz <- lin %*% t(matrix(c(
    0.4124564, 0.3575761, 0.1804375,
    0.2126729, 0.7151522, 0.0721750,
    0.0193339, 0.1191920, 0.9503041
  ), nrow = 3, byrow = TRUE))
  xyz <- sweep(xyz, 2, c(0.95047, 1, 1.08883), "/")
  eps <- (6 / 29)^3
  f <- ifelse(xyz > eps, xyz^(1 / 3), xyz / (3 * (6 / 29)^2) + 4 / 29)
  data.frame(
    L = 116 * f[, 2] - 16,
    a = 500 * (f[, 1] - f[, 2]),
    b = 200 * (f[, 2] - f[, 3]),
    check.names = FALSE
  )
}

lab <- srgb_to_lab_d65(df$R_primary, df$G_primary, df$B_primary)
df$L  <- lab$L
df$a  <- lab$a
df$b  <- lab$b
df$C  <- sqrt(df$a^2 + df$b^2)
df$Lm <- -df$L

hsv_mat <- t(grDevices::rgb2hsv(
  df$R_primary / 255,
  df$G_primary / 255,
  df$B_primary / 255
))
colnames(hsv_mat) <- c("H", "S", "V")
df <- cbind(df, as.data.frame(hsv_mat))
df$Sm <- df$S

df$cfa_qc_eligible <- TRUE
if ("qc_status" %in% names(df)) {
  df$cfa_qc_eligible <- df$cfa_qc_eligible & tolower(trimws(df$qc_status)) != "fail"
}
if ("duplicate_image_sha256" %in% names(df)) {
  duplicate_flag <- tolower(trimws(as.character(df$duplicate_image_sha256))) %in% c("1", "true", "yes", "y")
  df$cfa_qc_eligible <- df$cfa_qc_eligible & !duplicate_flag
}
if ("possible_overexposure" %in% names(df)) {
  over_flag <- tolower(trimws(as.character(df$possible_overexposure))) %in% c("1", "true", "yes", "y")
  df$cfa_qc_eligible <- df$cfa_qc_eligible & !over_flag
}
if ("mask_pixels" %in% names(df)) {
  df$cfa_qc_eligible <- df$cfa_qc_eligible & (is.na(df$mask_pixels) | as.numeric(df$mask_pixels) >= 50)
}
if ("mask_fraction_visible" %in% names(df)) {
  df$cfa_qc_eligible <- df$cfa_qc_eligible & (is.na(df$mask_fraction_visible) | as.numeric(df$mask_fraction_visible) > 0)
}

df$col_hex <- grDevices::rgb(
  pmin(pmax(round(df$R_primary), 0), 255),
  pmin(pmax(round(df$G_primary), 0), 255),
  pmin(pmax(round(df$B_primary), 0), 255),
  maxColorValue = 255
)

cat(
  "✓ Loaded extraction-v2 colour features. n =", nrow(df),
  "; CFA-QC eligible =", sum(df$cfa_qc_eligible, na.rm = TRUE), "\n"
)
```

## Environment

```{r public-environment-download}
cache_root <- file.path(out_root, "public_environment_cache")
chelsa_dir <- file.path(cache_root, "CHELSA_v2.1")
soil_dir   <- file.path(cache_root, "SoilGrids_2.0")
dem_dir    <- file.path(cache_root, "elevation")
bee_dir    <- file.path(out_root, "bombus_sdm", "predictions") # rebuilt below from public GBIF + public 30s predictors
human_dir  <- file.path(cache_root, "WorldPop_2020")
for (d in c(chelsa_dir, soil_dir, dem_dir, human_dir, bee_dir)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

df <- df %>% mutate(.row_id = row_number())
coord_ok <- is.finite(as.numeric(df$longitude)) & is.finite(as.numeric(df$latitude))
coords <- df[coord_ok, c(".row_id", "longitude", "latitude")]
coords$longitude <- as.numeric(coords$longitude)
coords$latitude  <- as.numeric(coords$latitude)
stopifnot(!anyDuplicated(coords$.row_id), nrow(coords) > 0)

bbox <- terra::ext(
  min(coords$longitude) - 0.15, max(coords$longitude) + 0.15,
  min(coords$latitude)  - 0.15, max(coords$latitude)  + 0.15
)

safe_download <- function(url, dest, min_bytes = 10000) {
  if (file.exists(dest) && file.info(dest)$size >= min_bytes) return(dest)
  dir.create(dirname(dest), recursive = TRUE, showWarnings = FALSE)
  tmp <- paste0(dest, ".part")
  unlink(tmp, force = TRUE)
  on.exit(unlink(tmp, force = TRUE), add = TRUE)

  errors <- character()

  if (requireNamespace("curl", quietly = TRUE)) {
    got <- try(
      curl::curl_download(
        url,
        tmp,
        quiet = FALSE,
        mode = "wb",
        handle = curl::new_handle(
          followlocation = TRUE,
          connecttimeout = 60,
          timeout = 0,
          ipresolve = 4L,
          useragent = "Campanula-thesis/1.0"
        )
      ),
      silent = TRUE
    )
    if (inherits(got, "try-error")) {
      errors <- c(errors, paste0("curl package: ", conditionMessage(attr(got, "condition"))))
    }
  } else {
    errors <- c(errors, "curl package: package not installed")
  }

  if (!file.exists(tmp) || file.info(tmp)$size < min_bytes) {
    unlink(tmp, force = TRUE)
    curl_bin <- Sys.which("curl")
    if (nzchar(curl_bin)) {
      status <- suppressWarnings(system2(
        curl_bin,
        c("-L", "--fail", "--retry", "5", "--retry-all-errors",
          "--connect-timeout", "60", "-4", "-o", shQuote(tmp), shQuote(url))
      ))
      if (!identical(status, 0L)) errors <- c(errors, paste0("system curl exit status: ", status))
    } else {
      errors <- c(errors, "system curl: executable not found")
    }
  }

  if (!file.exists(tmp) || file.info(tmp)$size < min_bytes) {
    unlink(tmp, force = TRUE)
    wget_bin <- Sys.which("wget")
    if (nzchar(wget_bin)) {
      status <- suppressWarnings(system2(
        wget_bin,
        c("--continue", "--tries=5", "--timeout=60", "-O", shQuote(tmp), shQuote(url))
      ))
      if (!identical(status, 0L)) errors <- c(errors, paste0("wget exit status: ", status))
    } else {
      errors <- c(errors, "wget: executable not found")
    }
  }

  if (!file.exists(tmp) || file.info(tmp)$size < min_bytes) {
    stop(
      "Download failed or incomplete: ", url, "\n",
      paste(errors, collapse = "\n"),
      "\nOpen the URL in a browser. If the browser also fails, the CHELSA host is unreachable from this network."
    )
  }
  if (!file.rename(tmp, dest)) stop("Could not move completed download to: ", dest)
  dest
}

crop_cache <- function(src, name, out_dir, extent = bbox) {
  out <- file.path(out_dir, paste0(name, "_Japan_crop.tif"))
  if (!file.exists(out)) {
    r <- terra::rast(src)
    r <- terra::crop(x = r, y = extent, snap = "out")
    names(r) <- name
    terra::writeRaster(x = r, filename = out, overwrite = TRUE, gdal = c("COMPRESS=DEFLATE"))
  }
  out
}

source_manifest <- tibble::tribble(
  ~variable, ~provider, ~version_period, ~unit_note, ~role,
  "bio5", "CHELSA", "v2.1; 1981-2010", "CHELSA native units; rescaled below", "temperature",
  "bio10", "CHELSA", "v2.1; 1981-2010", "CHELSA native units; rescaled below", "temperature",
  "bio12", "CHELSA", "v2.1; 1981-2010", "kg m-2 yr-1 (numerically mm yr-1)", "moisture",
  "bio14", "CHELSA", "v2.1; 1981-2010", "kg m-2 month-1", "moisture",
  "bio15", "CHELSA", "v2.1; 1981-2010", "coefficient of variation", "moisture seasonality",
  "CMI", "CHELSA", "v2.1; 1981-2010 climatological mean", "mean climatic moisture index", "AI replacement",
  "bdod/cfvo/nitrogen/ocd/soc/phh2o/sand/silt", "SoilGrids", "2.0; 0-5 cm mean", "property-specific mapped integer units; converted below", "soil",
  "elevation", "WorldClim elevation", "v2.1; 30 arc-second", "m", "topography",
  "population_count", "WorldPop", "2020; 1 km", "people per raster cell", "anthropogenic"
)
write_csv(source_manifest, file.path(cache_root, "environment_source_manifest.csv"))

chelsa_unil_base <- "https://os.unil.cloud.switch.ch/chelsa02/chelsa/global"

as_vsicurl <- function(url) {
  if (startsWith(url, "/vsi")) url else paste0("/vsicurl/", url)
}

open_remote_30s <- function(url) {
  old <- Sys.getenv(c(
    "GDAL_DISABLE_READDIR_ON_OPEN", "CPL_VSIL_CURL_ALLOWED_EXTENSIONS",
    "GDAL_HTTP_MULTIRANGE", "GDAL_HTTP_TIMEOUT", "GDAL_HTTP_MAX_RETRY",
    "GDAL_HTTP_RETRY_DELAY", "VSI_CACHE", "VSI_CACHE_SIZE"
  ), unset = NA_character_)
  on.exit({
    for (nm in names(old)) {
      if (is.na(old[[nm]])) {
        Sys.unsetenv(nm)
      } else {
        do.call(Sys.setenv, stats::setNames(list(old[[nm]]), nm))
      }
    }
  }, add = TRUE)
  Sys.setenv(
    GDAL_DISABLE_READDIR_ON_OPEN = "EMPTY_DIR",
    CPL_VSIL_CURL_ALLOWED_EXTENSIONS = ".tif,.tiff",
    GDAL_HTTP_MULTIRANGE = "YES",
    GDAL_HTTP_TIMEOUT = "120",
    GDAL_HTTP_MAX_RETRY = "4",
    GDAL_HTTP_RETRY_DELAY = "3",
    VSI_CACHE = "TRUE",
    VSI_CACHE_SIZE = "50000000"
  )
  r <- terra::rast(as_vsicurl(url))
  rs <- terra::res(r)
  target <- 1 / 120 # 30 arc-second in decimal degrees
  if (terra::is.lonlat(r) && any(abs(rs - target) > 1e-5)) {
    stop("CHELSA raster is not native 30 arc-second: ", url,
         " (resolution=", paste(signif(rs, 8), collapse = " x "), ")")
  }
  r
}

crop_remote_30s <- function(urls, stable_name, out_dir) {
  out <- file.path(out_dir, paste0(stable_name, "_Japan_crop_30s.tif"))
  if (file.exists(out) && file.info(out)$size > 10000) return(out)
  errors <- character()
  for (url in unique(urls)) {
    rr <- try(open_remote_30s(url), silent = TRUE)
    if (inherits(rr, "try-error")) {
      errors <- c(errors, paste0(url, " -> ", conditionMessage(attr(rr, "condition"))))
      next
    }
    ans <- try({
      if (!terra::is.lonlat(rr)) {
        stop("Expected a lon/lat CHELSA raster, got CRS: ", terra::crs(rr))
      }
      x <- terra::crop(x = rr, y = bbox, snap = "out")
      names(x) <- stable_name
      terra::writeRaster(
        x = x,
        filename = out,
        overwrite = TRUE,
        gdal = c("COMPRESS=DEFLATE", "TILED=YES")
      )
      out
    }, silent = TRUE)
    if (!inherits(ans, "try-error")) {
      message("CHELSA 30s source: ", url)
      return(ans)
    }
    errors <- c(errors, paste0(url, " -> ", conditionMessage(attr(ans, "condition"))))
  }
  stop("Could not open/crop CHELSA 30s layer ", stable_name, "\n",
       paste(errors, collapse = "\n"))
}

chelsa_bio <- c(bio4 = 4L, bio5 = 5L, bio6 = 6L, bio10 = 10L, bio12 = 12L, bio13 = 13L, bio14 = 14L, bio15 = 15L)
chelsa_files <- setNames(character(length(chelsa_bio)), names(chelsa_bio))
for (nm in names(chelsa_bio)) {
  xx <- sprintf("%02d", chelsa_bio[[nm]])
  fn <- sprintf("CHELSA_bio%s_1981-2010_V.2.1.tif", xx)
  url <- sprintf("%s/bioclim/bio%s/1981-2010/%s", chelsa_unil_base, xx, fn)
  chelsa_files[[nm]] <- crop_remote_30s(url, nm, chelsa_dir)
}

gdd5_fn <- "CHELSA_gdd5_1981-2010_V.2.1.tif"
gdd5_url <- sprintf("%s/bioclim/gdd5/1981-2010/%s", chelsa_unil_base, gdd5_fn)
chelsa_files[["gdd5"]] <- crop_remote_30s(gdd5_url, "gdd5", chelsa_dir)

cmi_fn <- "CHELSA_cmimean_1981-2010_V.2.1.tif"
cmi_url <- sprintf(
  "%s/bioclim/cmimean/1981-2010/%s",
  chelsa_unil_base,
  cmi_fn
)
chelsa_files[["CMI"]] <- crop_remote_30s(cmi_url, "CMI", chelsa_dir)

rsds_url <- paste0(
  "https://os.zhdk.cloud.switch.ch/chelsav2/GLOBAL/",
  "climatologies/1981-2010/bio/",
  "CHELSA_rsds_1981-2010_mean_V.2.1.tif"
)
chelsa_files[["RSDS"]] <- crop_remote_30s(
  urls = rsds_url,
  stable_name = "RSDS",
  out_dir = chelsa_dir
)

soil_registry <- tibble::tribble(
  ~variable,   ~map_name,   ~coverage_id,              ~scale_factor,
  "bdod",      "bdod",      "bdod_0-5cm_mean",         0.01,
  "cfvo",      "cfvo",      "cfvo_0-5cm_mean",         0.1,
  "nitrogen",  "nitrogen",  "nitrogen_0-5cm_mean",     0.01,
  "ocd",       "ocd",       "ocd_0-5cm_mean",          0.1,
  "soc",       "soc",       "soc_0-5cm_mean",          0.1,
  "phh2o",     "phh2o",     "phh2o_0-5cm_mean",        0.1,
  "sand",      "sand",      "sand_0-5cm_mean",         0.1,
  "silt",      "silt",      "silt_0-5cm_mean",         0.1
)

make_30s_template <- function(extent) {
  res30 <- 30 / 3600
  snap_down <- function(value, anchor) anchor + floor((value - anchor) / res30) * res30
  snap_up   <- function(value, anchor) anchor + ceiling((value - anchor) / res30) * res30
  terra::rast(
    xmin = snap_down(terra::xmin(extent), -180),
    xmax = snap_up(terra::xmax(extent), -180),
    ymin = snap_down(terra::ymin(extent), -90),
    ymax = snap_up(terra::ymax(extent), -90),
    resolution = res30,
    crs = "EPSG:4326"
  )
}
soil_template_30s <- make_30s_template(bbox)

soilgrids_wcs_url <- function(map_name, coverage_id, extent,
                              width = 11000L, height = 5400L) {
  roi_ll <- terra::as.polygons(
    terra::ext(
      terra::xmin(extent), terra::xmax(extent),
      terra::ymin(extent), terra::ymax(extent)
    ),
    crs = "EPSG:4326"
  )
  roi_sg <- terra::project(x = roi_ll, y = "ESRI:54052")
  projected <- as.vector(terra::ext(roi_sg))

  paste0(
    "https://maps.isric.org/mapserv?map=/map/", map_name, ".map",
    "&SERVICE=WCS&VERSION=2.0.1&REQUEST=GetCoverage",
    "&COVERAGEID=", utils::URLencode(coverage_id, reserved = TRUE),
    "&FORMAT=GEOTIFF_INT16",
    "&SUBSET=x(", format(projected[[1]], scientific = FALSE, digits = 17L),
    ",", format(projected[[2]], scientific = FALSE, digits = 17L), ")",
    "&SUBSET=y(", format(projected[[3]], scientific = FALSE, digits = 17L),
    ",", format(projected[[4]], scientific = FALSE, digits = 17L), ")",
    "&SCALESIZE=x(", as.integer(width), "),y(", as.integer(height), ")",
    "&INTERPOLATION=NEAREST"
  )
}

cache_soilgrids_30s <- function(variable, map_name, coverage_id, scale_factor,
                                out_dir, extent = bbox, template = soil_template_30s) {
  native_out <- file.path(out_dir, paste0(variable, "_0-5cm_mean_WCS_native.tif"))
  final_out  <- file.path(out_dir, paste0(variable, "_0-5cm_mean_30s.tif"))
  if (file.exists(final_out) && file.info(final_out)$size > 10000) return(final_out)

  request <- soilgrids_wcs_url(
    map_name = map_name,
    coverage_id = coverage_id,
    extent = extent
  )

  old <- Sys.getenv(c("GDAL_HTTP_TIMEOUT", "GDAL_HTTP_MAX_RETRY"), unset = NA_character_)
  on.exit({
    for (nm in names(old)) {
      if (is.na(old[[nm]])) {
        Sys.unsetenv(nm)
      } else {
        do.call(Sys.setenv, stats::setNames(list(old[[nm]]), nm))
      }
    }
  }, add = TRUE)
  Sys.setenv(GDAL_HTTP_TIMEOUT = "1800", GDAL_HTTP_MAX_RETRY = "3")

  if (!file.exists(native_out) || file.info(native_out)$size <= 10000) {
    native <- terra::rast(request)
    if (terra::nlyr(native) != 1L) {
      stop("Expected one SoilGrids layer for ", variable)
    }
    if (!nzchar(terra::crs(native))) terra::crs(native) <- "ESRI:54052"
    terra::writeRaster(
      x = native,
      filename = native_out,
      overwrite = TRUE,
      gdal = c("TILED=YES", "COMPRESS=DEFLATE", "PREDICTOR=2")
    )
  }

  native <- terra::rast(native_out)
  native <- native * as.numeric(scale_factor)
  aligned <- terra::project(
    x = native,
    y = template,
    method = "bilinear",
    mask = TRUE
  )
  names(aligned) <- variable
  terra::writeRaster(
    x = aligned,
    filename = final_out,
    overwrite = TRUE,
    gdal = c("TILED=YES", "COMPRESS=DEFLATE", "PREDICTOR=2")
  )
  final_out
}

soil_files <- setNames(character(nrow(soil_registry)), soil_registry$variable)
for (i in seq_len(nrow(soil_registry))) {
  row <- soil_registry[i, ]
  soil_files[[row$variable]] <- cache_soilgrids_30s(
    variable = row$variable,
    map_name = row$map_name,
    coverage_id = row$coverage_id,
    scale_factor = row$scale_factor,
    out_dir = soil_dir
  )
}

worldpop_url <- paste0(
  "https://data.worldpop.org/GIS/Population/Global_2000_2020_1km/",
  "2020/JPN/jpn_ppp_2020_1km_Aggregated.tif"
)
worldpop_raw <- safe_download(
  worldpop_url,
  file.path(human_dir, "jpn_ppp_2020_1km_Aggregated.tif"),
  min_bytes = 100000
)
worldpop_file <- crop_cache(worldpop_raw, "population_count", human_dir)

dem_zip <- safe_download(
  "https://geodata.ucdavis.edu/climate/worldclim/2_1/base/wc2.1_30s_elev.zip",
  file.path(dem_dir, "wc2.1_30s_elev.zip"),
  min_bytes = 1000000
)
dem_unzip_dir <- file.path(dem_dir, "wc2.1_30s_elev")
dir.create(dem_unzip_dir, recursive = TRUE, showWarnings = FALSE)
dem_candidates <- list.files(dem_unzip_dir, pattern = "\\.tif$", full.names = TRUE)
if (!length(dem_candidates)) {
  utils::unzip(dem_zip, exdir = dem_unzip_dir)
  dem_candidates <- list.files(dem_unzip_dir, pattern = "\\.tif$", full.names = TRUE)
}
if (!length(dem_candidates)) stop("Elevation raster was not found after extracting: ", dem_zip)
dem_file <- crop_cache(dem_candidates[[1]], "elevation", dem_dir)

public_files <- c(chelsa_files, soil_files, elevation = dem_file, population_count = worldpop_file)
```

## Bombus SDM

```{r rebuild_bombus_sdm}

message("RUNNING PIPELINE VERSION: FINAL_PUBLICATION_CLEAN_20KM")
REBUILD_BOMBUS_SDM <- isTRUE(getOption("campanula.rebuild_bombus_sdm", TRUE))
BOMBUS_FORCE_GBIF_REFRESH <- isTRUE(getOption("campanula.refresh_gbif", FALSE))
BOMBUS_MAX_COORD_UNCERTAINTY_M <- 10000
BOMBUS_BACKGROUND_N <- 10000
BOMBUS_SEED <- 42
set.seed(BOMBUS_SEED)

bombus_root <- file.path(out_root, "bombus_sdm")
bombus_occ_dir <- file.path(bombus_root, "occurrences")
bombus_model_dir <- file.path(bombus_root, "models")
bombus_eval_dir <- file.path(bombus_root, "evaluation")
for (d in c(bombus_root, bombus_occ_dir, bombus_model_dir, bombus_eval_dir, bee_dir)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}

bombus_species <- tibble::tribble(
  ~species, ~short,
  "Bombus ardens", "ardens",
  "Bombus beaticola", "beaticola",
  "Bombus consobrinus", "consobrinus",
  "Bombus diversus", "diversus",
  "Bombus honshuensis", "honshuensis"
)

required_sdm_pkgs <- c("rgbif", "ENMeval", "maxnet", "usdm")

install_missing_cran_packages <- function(packages) {
  missing <- packages[
    !vapply(packages, requireNamespace, logical(1), quietly = TRUE)
  ]
  if (!length(missing)) return(invisible(character()))

  message("Installing missing Bombus-SDM packages: ", paste(missing, collapse = ", "))
  repos <- getOption("repos")
  if (is.null(repos) || identical(unname(repos[["CRAN"]]), "@CRAN@")) {
    repos <- c(CRAN = "https://cloud.r-project.org")
  }

  tryCatch(
    utils::install.packages(
      missing,
      repos = repos,
      dependencies = TRUE,
      Ncpus = max(1L, parallel::detectCores(logical = FALSE) - 1L)
    ),
    error = function(e) {
      stop(
        "Automatic installation of Bombus-SDM packages failed: ",
        conditionMessage(e),
        "\nInstall manually with:\n",
        'install.packages(c(\"rgbif\", \"ENMeval\", \"maxnet\", \"usdm\"), dependencies = TRUE)',
        call. = FALSE
      )
    }
  )

  still_missing <- packages[
    !vapply(packages, requireNamespace, logical(1), quietly = TRUE)
  ]
  if (length(still_missing)) {
    stop(
      "Packages remain unavailable after installation: ",
      paste(still_missing, collapse = ", "),
      ". Restart R and knit again. If usdm alone fails, install it separately before rerunning.",
      call. = FALSE
    )
  }
  invisible(missing)
}

if (REBUILD_BOMBUS_SDM) {
  install_missing_cran_packages(required_sdm_pkgs)
}

`%||%` <- function(x, y) {
  if (is.null(x) || !length(x) || all(is.na(x))) y else x
}

extract_gbif_taxon_key <- function(x) {
  candidates <- list(
    x$usageKey,
    x$taxonKey,
    x$key,
    if (is.list(x$usage)) x$usage$key else NULL,
    if (is.list(x$acceptedUsage)) x$acceptedUsage$key else NULL
  )
  for (candidate in candidates) {
    if (!is.null(candidate) && length(candidate) == 1L && !is.na(candidate)) {
      candidate <- as.character(candidate)
      if (nzchar(candidate)) return(candidate)
    }
  }
  NA_character_
}

resolve_gbif_taxon <- function(scientific_name) {
  matched <- tryCatch(
    rgbif::name_backbone(
      name = scientific_name,
      kingdom = "Animalia",
      rank = "species",
      strict = FALSE
    ),
    error = function(e) NULL
  )
  key <- if (is.null(matched)) NA_character_ else extract_gbif_taxon_key(matched)

  if (is.na(key)) {
    suggested <- tryCatch(
      rgbif::name_suggest(q = scientific_name, rank = "species", limit = 50)$data,
      error = function(e) NULL
    )
    if (!is.null(suggested) && nrow(suggested)) {
      candidate_name <- tolower(trimws(scientific_name))
      name_columns <- intersect(
        c("canonicalName", "scientificName", "species"),
        names(suggested)
      )
      exact <- rep(FALSE, nrow(suggested))
      for (column in name_columns) {
        exact <- exact | tolower(trimws(as.character(suggested[[column]]))) == candidate_name
      }
      suggested <- suggested[exact, , drop = FALSE]
      if (nrow(suggested)) {
        for (column in intersect(c("key", "usageKey", "taxonKey"), names(suggested))) {
          value <- as.character(suggested[[column]][1])
          if (!is.na(value) && nzchar(value)) {
            key <- value
            break
          }
        }
      }
    }
  }

  if (is.na(key)) {
    stop(
      "GBIF taxon key unresolved after backbone and exact-suggestion lookup: ",
      scientific_name,
      ". This can occur during GBIF taxonomy reprocessing. Retry later or use a frozen occurrence cache.",
      call. = FALSE
    )
  }

  list(key = key, match = matched)
}

fetch_gbif_species <- function(scientific_name, cache_csv, refresh = FALSE) {
  if (file.exists(cache_csv) && !refresh) {
    return(readr::read_csv(cache_csv, show_col_types = FALSE))
  }

  taxon <- resolve_gbif_taxon(scientific_name)
  key <- taxon$key
  message("GBIF taxon resolved: ", scientific_name, " -> ", key)

  page_size <- 300L
  start <- 0L
  pages <- list()
  repeat {
    ans <- rgbif::occ_search(
      taxonKey = key,
      country = "JP",
      hasCoordinate = TRUE,
      occurrenceStatus = "present",
      limit = page_size,
      start = start,
      fields = c(
        "key", "scientificName", "decimalLongitude", "decimalLatitude",
        "coordinateUncertaintyInMeters", "basisOfRecord", "year",
        "datasetKey", "publishingOrgKey", "issues"
      )
    )
    if (!nrow(ans$data)) break
    pages[[length(pages) + 1L]] <- ans$data
    start <- start + nrow(ans$data)
    if (start >= ans$meta$count || nrow(ans$data) < page_size) break
  }

  out <- dplyr::bind_rows(pages)
  if (!nrow(out)) stop("No GBIF records returned for ", scientific_name)

  accepted_basis <- c(
    "HUMAN_OBSERVATION", "OBSERVATION", "PRESERVED_SPECIMEN",
    "MATERIAL_SAMPLE", "MACHINE_OBSERVATION"
  )
  out <- out |>
    dplyr::mutate(
      decimalLongitude = as.numeric(decimalLongitude),
      decimalLatitude = as.numeric(decimalLatitude),
      coordinateUncertaintyInMeters = as.numeric(coordinateUncertaintyInMeters)
    ) |>
    dplyr::filter(
      is.finite(decimalLongitude), is.finite(decimalLatitude),
      decimalLongitude >= 120, decimalLongitude <= 150,
      decimalLatitude >= 20, decimalLatitude <= 50,
      is.na(coordinateUncertaintyInMeters) |
        coordinateUncertaintyInMeters <= BOMBUS_MAX_COORD_UNCERTAINTY_M,
      is.na(basisOfRecord) | basisOfRecord %in% accepted_basis
    ) |>
    dplyr::distinct(key, .keep_all = TRUE)

  readr::write_csv(out, cache_csv)
  out
}

as_terra_xy <- function(x, lon = "lon", lat = "lat") {
  if (!all(c(lon, lat) %in% names(x))) {
    stop("Coordinate columns missing: ", lon, ", ", lat, call. = FALSE)
  }
  lon_v <- suppressWarnings(as.numeric(unlist(x[[lon]], use.names = FALSE)))
  lat_v <- suppressWarnings(as.numeric(unlist(x[[lat]], use.names = FALSE)))
  if (length(lon_v) != nrow(x) || length(lat_v) != nrow(x)) {
    stop("Coordinate columns are not scalar numeric vectors.", call. = FALSE)
  }
  xy <- cbind(lon = lon_v, lat = lat_v)
  storage.mode(xy) <- "double"
  xy
}

select_best_enmeval <- function(results) {
  idx <- seq_len(nrow(results))
  if ("cbi.val.avg" %in% names(results) && any(is.finite(results$cbi.val.avg))) {
    best_cbi <- max(results$cbi.val.avg, na.rm = TRUE)
    idx <- idx[is.finite(results$cbi.val.avg[idx]) &
                 results$cbi.val.avg[idx] >= best_cbi - 0.02]
  }
  if ("or.10p.val.avg" %in% names(results) && length(idx) > 1L &&
      any(is.finite(results$or.10p.val.avg[idx]))) {
    best_or <- min(results$or.10p.val.avg[idx], na.rm = TRUE)
    idx <- idx[is.finite(results$or.10p.val.avg[idx]) &
                 results$or.10p.val.avg[idx] <= best_or + 0.02]
  }
  if ("AICc" %in% names(results) && length(idx) > 1L &&
      any(is.finite(results$AICc[idx]))) {
    idx <- idx[which.min(results$AICc[idx])]
  }
  idx[[1]]
}

predict_maxnet_cloglog <- function(env_rast, model) {
  terra::predict(
    object = env_rast,
    model = model,
    fun = function(m, newdata, ...) {
      as.numeric(stats::predict(
        m,
        newdata = as.data.frame(newdata),
        type = "cloglog"
      ))
    },
    na.rm = TRUE
  )
}

if (REBUILD_BOMBUS_SDM) {
  sdm_base_files <- c(
    bio4 = chelsa_files[["bio4"]],
    bio6 = chelsa_files[["bio6"]],
    bio13 = chelsa_files[["bio13"]],
    CMI = chelsa_files[["CMI"]],
    RSDS = chelsa_files[["RSDS"]],
    bdod = soil_files[["bdod"]],
    cfvo = soil_files[["cfvo"]],
    sand = soil_files[["sand"]],
    silt = soil_files[["silt"]],
    nitrogen = soil_files[["nitrogen"]],
    soil_pH = soil_files[["phh2o"]],
    soc = soil_files[["soc"]],
    elevation = dem_file
  )
  stopifnot(all(file.exists(sdm_base_files)))

  sdm_env <- terra::rast(unname(sdm_base_files))
  names(sdm_env) <- names(sdm_base_files)

  dem_sdm <- terra::rast(dem_file)
  sdm_slope <- terra::terrain(dem_sdm, v = "slope", unit = "degrees")
  sdm_rough <- terra::terrain(dem_sdm, v = "roughness")
  names(sdm_slope) <- "slope"
  names(sdm_rough) <- "roughness"
  sdm_slope <- terra::resample(sdm_slope, sdm_env[[1]], method = "bilinear")
  sdm_rough <- terra::resample(sdm_rough, sdm_env[[1]], method = "bilinear")
  sdm_env <- c(sdm_env, sdm_slope, sdm_rough)

  occurrence_list <- list()
  occurrence_manifest <- list()
  for (i in seq_len(nrow(bombus_species))) {
    sp <- bombus_species$species[[i]]
    sh <- bombus_species$short[[i]]
    cache_csv <- file.path(bombus_occ_dir, paste0(sh, "_gbif.csv"))
    occ_i <- fetch_gbif_species(sp, cache_csv, BOMBUS_FORCE_GBIF_REFRESH)
    occurrence_list[[sh]] <- occ_i
    occurrence_manifest[[sh]] <- tibble::tibble(
      species = sp,
      short = sh,
      n_raw_cached = nrow(occ_i),
      accessed_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
      cache_csv = normalizePath(cache_csv, winslash = "/", mustWork = TRUE),
      dataset_keys = paste(sort(unique(stats::na.omit(occ_i$datasetKey))), collapse = ";")
    )
  }
  readr::write_csv(
    dplyr::bind_rows(occurrence_manifest),
    file.path(bombus_root, "gbif_occurrence_manifest.csv")
  )

  pooled <- dplyr::bind_rows(
    lapply(names(occurrence_list), function(sh) {
      occurrence_list[[sh]] |>
        dplyr::transmute(
          short = sh,
          lon = decimalLongitude,
          lat = decimalLatitude
        )
    })
  )
  pooled_cells <- terra::cellFromXY(sdm_env[[1]], as_terra_xy(pooled))
  pooled <- pooled[!is.na(pooled_cells) & !duplicated(pooled_cells), , drop = FALSE]
  pooled_cells <- terra::cellFromXY(sdm_env[[1]], as_terra_xy(pooled))

  make_species_M <- function(occ_df, species_short,
                             min_buffer_km = 100,
                             max_buffer_km = 300,
                             range_fraction = 0.15) {
    pts <- sf::st_as_sf(occ_df, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
    pts_ea <- sf::st_transform(pts, 6933)
    bb <- sf::st_bbox(pts_ea)
    range_diag_m <- sqrt((bb[["xmax"]] - bb[["xmin"]])^2 +
                         (bb[["ymax"]] - bb[["ymin"]])^2)
    buffer_km <- max(min_buffer_km,
                     min(max_buffer_km, range_fraction * range_diag_m / 1000))
    m_ea <- sf::st_buffer(sf::st_convex_hull(sf::st_union(pts_ea)), dist = buffer_km * 1000)
    m_ll <- sf::st_transform(sf::st_as_sf(m_ea), 4326)
    m_ll$species <- species_short
    m_ll$buffer_km <- buffer_km
    m_ll$n_occurrence_cells <- nrow(occ_df)
    m_ll
  }

  sdm_metrics <- list()
  sdm_warning_log <- list()
  for (i in seq_len(nrow(bombus_species))) {
    sp <- bombus_species$species[[i]]
    sh <- bombus_species$short[[i]]
    occ_i <- occurrence_list[[sh]] |>
      dplyr::transmute(lon = decimalLongitude, lat = decimalLatitude) |>
      dplyr::filter(is.finite(lon), is.finite(lat))

    occ_cells <- terra::cellFromXY(sdm_env[[1]], as_terra_xy(occ_i))
    occ_i <- occ_i[!is.na(occ_cells) & !duplicated(occ_cells), , drop = FALSE]
    if (nrow(occ_i) < 20L) stop("Too few one-per-cell records for ", sp, ": ", nrow(occ_i))

    species_M_sf <- make_species_M(occ_i, sh)
    species_M_file <- file.path(bombus_model_dir, paste0(sh, "_accessible_area_M.gpkg"))
    if (file.exists(species_M_file)) unlink(species_M_file)
    sf::st_write(species_M_sf, species_M_file, quiet = TRUE)
    species_M <- terra::vect(species_M_sf)

    species_env <- terra::crop(sdm_env, species_M, snap = "out")
    species_env <- terra::mask(species_env, species_M)
    if (terra::ncell(species_env) < 1000L) {
      stop("Accessible area is too small for ", sp, ". Check occurrence coordinates.")
    }

    domain_sample <- terra::spatSample(
      species_env,
      size = min(20000, terra::ncell(species_env)),
      method = "random",
      na.rm = TRUE,
      values = TRUE,
      as.df = TRUE
    )
    domain_sample <- domain_sample[stats::complete.cases(domain_sample), , drop = FALSE]
    if (nrow(domain_sample) < 200L) stop("Too few complete environmental cells in M for ", sp)
    vif_fit <- usdm::vifstep(domain_sample, th = 10)
    kept_sdm_vars <- as.character(vif_fit@results$Variables)
    if (!length(kept_sdm_vars)) stop("No Bombus SDM predictor survived VIF filtering for ", sp)
    species_env_kept <- species_env[[kept_sdm_vars]]
    readr::write_csv(
      as.data.frame(vif_fit@results),
      file.path(bombus_eval_dir, paste0(sh, "_sdm_predictor_vif.csv"))
    )

    pooled_cells_i <- terra::cellFromXY(species_env_kept[[1]], as_terra_xy(pooled))
    bg_pool <- pooled[!is.na(pooled_cells_i), c("lon", "lat"), drop = FALSE]
    occ_cells_i <- terra::cellFromXY(species_env_kept[[1]], as_terra_xy(occ_i))
    bg_cells_i <- terra::cellFromXY(species_env_kept[[1]], as_terra_xy(bg_pool))
    bg_pool <- bg_pool[!is.na(bg_cells_i) & !bg_cells_i %in% occ_cells_i & !duplicated(bg_cells_i), , drop = FALSE]
    if (nrow(bg_pool) > BOMBUS_BACKGROUND_N) {
      bg_pool <- bg_pool[sample.int(nrow(bg_pool), BOMBUS_BACKGROUND_N), , drop = FALSE]
    }
    if (nrow(bg_pool) < 500L) {
      fallback <- terra::spatSample(
        species_env_kept,
        size = min(BOMBUS_BACKGROUND_N, terra::ncell(species_env_kept)),
        method = "random",
        na.rm = TRUE,
        values = FALSE,
        xy = TRUE
      )
      bg_pool <- as.data.frame(fallback)
      names(bg_pool) <- c("lon", "lat")
      fallback_cells <- terra::cellFromXY(species_env_kept[[1]], as_terra_xy(bg_pool))
      bg_pool <- bg_pool[!is.na(fallback_cells) & !fallback_cells %in% occ_cells_i & !duplicated(fallback_cells), , drop = FALSE]
    }
    if (nrow(bg_pool) < 500L) stop("Too few background cells inside M for ", sp)

    eval_capture <- capture_warnings(ENMeval::ENMevaluate(
      occs = as.data.frame(as_terra_xy(occ_i)),
      envs = species_env_kept,
      bg = as.data.frame(as_terra_xy(bg_pool)),
      algorithm = "maxnet",
      partitions = "block",
      tune.args = list(fc = c("L", "LQ", "LQH"), rm = 1:5),
      parallel = FALSE,
      quiet = TRUE
    ))
    eval_obj <- eval_capture$value
    if (length(eval_capture$warnings)) {
      sdm_warning_log[[sh]] <- tibble::tibble(species = sp, warning = eval_capture$warnings)
    }
    res <- ENMeval::eval.results(eval_obj)
    best_i <- select_best_enmeval(res)
    best_model <- ENMeval::eval.models(eval_obj)[[best_i]]

    pred <- predict_maxnet_cloglog(species_env_kept, best_model)
    pred <- terra::mask(pred, species_M)
    names(pred) <- sh
    pred_file <- file.path(bee_dir, paste0(sh, ".tif"))
    terra::writeRaster(pred, pred_file, overwrite = TRUE)

    saveRDS(eval_obj, file.path(bombus_model_dir, paste0(sh, "_ENMeval.rds")))
    readr::write_csv(res, file.path(bombus_eval_dir, paste0(sh, "_all_models.csv")))
    readr::write_csv(
      dplyr::bind_cols(
        tibble::tibble(species = sp, short = sh, selected_row = best_i),
        res[best_i, , drop = FALSE]
      ),
      file.path(bombus_eval_dir, paste0(sh, "_selected_model.csv"))
    )
    sdm_metrics[[sh]] <- dplyr::bind_cols(
      tibble::tibble(
        species = sp,
        short = sh,
        n_occurrence_cells = nrow(occ_i),
        n_background = nrow(bg_pool),
        accessible_area_definition = "buffered_MCP_in_EPSG6933",
        accessible_area_buffer_km = species_M_sf$buffer_km[[1]],
        prediction_outside_M = "NA_not_evaluated",
        output_scale = "maxnet_cloglog_relative_suitability_not_calibrated_presence_probability",
        predictors = paste(kept_sdm_vars, collapse = ";")
      ),
      res[best_i, , drop = FALSE]
    )
  }

  readr::write_csv(
    dplyr::bind_rows(sdm_metrics),
    file.path(bombus_root, "bombus_sdm_manifest.csv")
  )
  sdm_warnings <- dplyr::bind_rows(sdm_warning_log)
  readr::write_csv(sdm_warnings, file.path(bombus_root, "ENMeval_warnings.csv"))
  if (nrow(sdm_warnings)) print(sdm_warnings) else cat("No ENMeval warnings.\n")
}

expected_bee_files <- file.path(bee_dir, paste0(bombus_species$short, ".tif"))
if (any(!file.exists(expected_bee_files))) {
  stop(
    "Missing reproducible Bombus predictions: ",
    paste(expected_bee_files[!file.exists(expected_bee_files)], collapse = ", "),
    call. = FALSE
  )
}
```

## Extraction

```{r extract_public_predictors}
extract_files_by_row <- function(files, coords) {
  pts <- terra::vect(coords, geom = c("longitude", "latitude"), crs = "EPSG:4326", keepgeom = FALSE)
  vals <- lapply(files, function(f) {
    r <- terra::rast(f)
    p <- if (terra::same.crs(pts, r)) pts else terra::project(x = pts, y = terra::crs(r))
    terra::extract(r, p, ID = FALSE)[, 1]
  })
  out <- as.data.frame(vals)
  names(out) <- names(files)
  dplyr::bind_cols(tibble(.row_id = coords$.row_id), out)
}

env_vals <- extract_files_by_row(public_files, coords)

bee_tifs <- expected_bee_files
names(bee_tifs) <- bombus_species$short
stopifnot(all(file.exists(bee_tifs)))
bee_vals <- extract_files_by_row(bee_tifs, coords)

df <- df %>%
  left_join(env_vals, by = ".row_id") %>%
  left_join(bee_vals, by = ".row_id")


df2 <- df %>% filter(coord_ok[.row_id]) %>% distinct(longitude, latitude, .keep_all = TRUE)
```

## Topography

```{r dem-topography}
if (file.exists(dem_file)) {
  dem_r <- terra::rast(dem_file)

  slope_r  <- terra::terrain(dem_r, v = "slope",  unit = "radians")
  aspect_r <- terra::terrain(dem_r, v = "aspect", unit = "radians")
  tri_r    <- terra::terrain(dem_r, v = "TRI")
  tpi_r    <- terra::terrain(dem_r, v = "TPI")
  rough_r  <- terra::terrain(dem_r, v = "roughness")

  names(slope_r)  <- "slope"
  names(aspect_r) <- "aspect"
  names(tri_r)    <- "TRI"
  names(tpi_r)    <- "TPI"
  names(rough_r)  <- "roughness"

  topo_stack <- c(slope_r, aspect_r, tri_r, tpi_r, rough_r)

  pts <- terra::vect(
    df2 %>% dplyr::select(longitude, latitude),
    geom = c("longitude", "latitude"),
    crs  = "EPSG:4326"
  )

  pts_use <- if (!terra::same.crs(pts, topo_stack)) {
    terra::project(x = pts, y = terra::crs(topo_stack))
  } else {
    pts
  }

  topo_vals <- terra::extract(topo_stack, pts_use, ID = FALSE)
  df2 <- cbind(df2, topo_vals)

  df2$northness <- cos(df2$aspect)
  df2$eastness  <- sin(df2$aspect)

} else {
  cat("! DEM file not found:", dem_file, "\n")
}
```

## Environmental axes

```{r pca}
set.seed(42)

temp_vars <- intersect(c("bio5", "bio10", "gdd5"), names(df2))

Temperature_PC1 <- rep(NA_real_, nrow(df2))
temp_pc_obj <- NULL

if (length(temp_vars) >= 1) {
  Xt <- df2[, temp_vars, drop = FALSE]
  cc <- complete.cases(Xt)
  if (sum(cc) >= 3 && ncol(Xt) >= 2) {
    temp_pc_obj <- prcomp(Xt[cc, ], scale. = TRUE)
    Temperature_PC1[cc] <- temp_pc_obj$x[, 1]
  } else if (ncol(Xt) == 1) {
    Temperature_PC1 <- as.numeric(scale(Xt[[1]]))
  }
}
df2$Temperature_PC1 <- Temperature_PC1

precip_vars <- intersect(c("CMI", "bio12", "bio14"), names(df2))

precip_PC1 <- rep(NA_real_, nrow(df2))
precip_pc_obj <- NULL

if (length(precip_vars) >= 1) {
  Xp <- df2[, precip_vars, drop = FALSE]
  cc <- complete.cases(Xp)
  if (sum(cc) >= 3 && ncol(Xp) >= 2) {
    precip_pc_obj <- prcomp(Xp[cc, ], scale. = TRUE)
    precip_PC1[cc] <- precip_pc_obj$x[, 1]
  } else if (ncol(Xp) == 1) {
    precip_PC1 <- as.numeric(scale(Xp[[1]]))
  }
}
df2$precip_PC1 <- precip_PC1
stopifnot("CMI" %in% precip_vars, !any(grepl("(^|_)AI($|_)", precip_vars, ignore.case = TRUE)))
write_csv(tibble(variable = precip_vars), file.path(cmp_dir, "moisture_pca_variables.csv"))
write_csv(
  tibble::tribble(
    ~axis, ~source_variable, ~interpretation,
    "TemperatureSeasonality", "BIO4", "interannual climatological temperature variability",
    "PrecipSeasonality", "BIO15", "interannual climatological precipitation variability"
  ),
  file.path(cmp_dir, "seasonality_predictor_definitions.csv")
)
if (!("bio15" %in% names(df2))) stop("BIO15 is required as a separate precipitation-seasonality predictor.")
df2$PrecipSeasonality <- z(df2$bio15)
if (!("bio4" %in% names(df2))) stop("BIO4 is required as a separate temperature-seasonality predictor.")
df2$TemperatureSeasonality <- z(df2$bio4)

soil_phys_vars <- intersect(c("bdod", "cfvo", "sand", "silt"), names(df2))

soil_phys_PC1 <- rep(NA_real_, nrow(df2))
soil_phys_pc_obj <- NULL

if (length(soil_phys_vars) >= 1) {
  Xp <- df2[, soil_phys_vars, drop = FALSE]
  keep <- vapply(Xp, function(v) is.numeric(v) && sd(v, na.rm = TRUE) > 0, logical(1))
  Xp <- Xp[, keep, drop = FALSE]
  if (ncol(Xp) >= 2) {
    cc <- complete.cases(Xp)
    if (sum(cc) >= 3) {
      soil_phys_pc_obj <- prcomp(Xp[cc, ], scale. = TRUE)
      soil_phys_PC1[cc] <- soil_phys_pc_obj$x[, 1]
    }
  } else if (ncol(Xp) == 1) {
    soil_phys_PC1 <- as.numeric(scale(Xp[[1]]))
  }
}
df2$soil_phys_PC1 <- soil_phys_PC1

soil_nutrient_vars <- intersect(c("nitrogen", "ocd", "soc"), names(df2))

soil_nutrient_PC1 <- rep(NA_real_, nrow(df2))
soil_nutrient_pc_obj <- NULL

if (length(soil_nutrient_vars) >= 1) {
  Xn <- df2[, soil_nutrient_vars, drop = FALSE]
  keep <- vapply(Xn, function(v) is.numeric(v) && sd(v, na.rm = TRUE) > 0, logical(1))
  Xn <- Xn[, keep, drop = FALSE]
  if (ncol(Xn) >= 2) {
    cc <- complete.cases(Xn)
    if (sum(cc) >= 3) {
      soil_nutrient_pc_obj <- prcomp(Xn[cc, ], scale. = TRUE)
      soil_nutrient_PC1[cc] <- soil_nutrient_pc_obj$x[, 1]
    }
  } else if (ncol(Xn) == 1) {
    soil_nutrient_PC1 <- as.numeric(scale(Xn[[1]]))
  }
}
df2$soil_nutrient_PC1 <- soil_nutrient_PC1

soil_pH_var <- intersect("phh2o", names(df2))
if (length(soil_pH_var) >= 1) {
  df2$soil_pH <- df2[[soil_pH_var[1]]]
} else {
  df2$soil_pH <- NA_real_
}

topo_vars <- intersect(c("roughness", "slope", "TRI"), names(df2))

topo_PC1 <- rep(NA_real_, nrow(df2))
topo_pc_obj <- NULL

if (length(topo_vars) >= 2) {
  Xt <- df2[, topo_vars, drop = FALSE]
  keep <- vapply(Xt, function(v) is.numeric(v) && sd(v, na.rm = TRUE) > 0, logical(1))
  Xt <- Xt[, keep, drop = FALSE]
  if (ncol(Xt) >= 2) {
    cc <- complete.cases(Xt)
    if (sum(cc) >= 3) {
      topo_pc_obj <- prcomp(Xt[cc, ], scale. = TRUE)
      topo_PC1[cc] <- topo_pc_obj$x[, 1]
    }
  } else if (ncol(Xt) == 1) {
    topo_PC1 <- as.numeric(scale(Xt[[1]]))
  }
}
df2$topo_PC1 <- topo_PC1

```

## Pigmentation index

```{r cfa-pigment}
cfa_vars <- c("a", "Lm", "C", "Sm")
stopifnot(all(cfa_vars %in% names(df2)))

cfa_rows <- which(df2$cfa_qc_eligible & complete.cases(df2[, cfa_vars]))
if (length(cfa_rows) < 100L) {
  stop("Fewer than 100 QC-eligible complete records for upgraded CFA.", call. = FALSE)
}

cfa_data <- as.data.frame(scale(df2[cfa_rows, cfa_vars]))
names(cfa_data) <- paste0(cfa_vars, "_z")

model_pig_v2 <- '
  Pigment =~ a_z + Lm_z + C_z + Sm_z
  a_z ~~ C_z
'

fit_pig <- lavaan::cfa(
  model = model_pig_v2,
  data = cfa_data,
  estimator = "MLR",
  std.lv = TRUE,
  meanstructure = TRUE,
  missing = "fiml"
)

if (!isTRUE(lavaan::inspect(fit_pig, "converged"))) {
  stop("Upgraded pigmentation CFA did not converge.", call. = FALSE)
}

score_qc <- as.numeric(lavaan::lavPredict(fit_pig, type = "lv")[, "Pigment"])
loading_v2 <- lavaan::standardizedSolution(fit_pig)
loading_v2 <- loading_v2[loading_v2$op == "=~", c("lhs", "rhs", "est.std", "se", "pvalue")]
if (mean(loading_v2$est.std, na.rm = TRUE) < 0) score_qc <- -score_qc

df2$Pigment <- NA_real_
df2$Pigment[cfa_rows] <- score_qc

fit_measures <- lavaan::fitMeasures(
  fit_pig,
  c("chisq.scaled", "df.scaled", "pvalue.scaled", "cfi.scaled",
    "tli.scaled", "rmsea.scaled", "rmsea.ci.lower.scaled",
    "rmsea.ci.upper.scaled", "srmr", "aic", "bic")
)

std_sol <- lavaan::standardizedSolution(fit_pig)
lambda <- std_sol$est.std[std_sol$op == "=~"]
theta <- std_sol$est.std[std_sol$op == "~~" & std_sol$lhs == std_sol$rhs &
                           std_sol$lhs %in% paste0(cfa_vars, "_z")]
omega_cfa <- (sum(lambda)^2) / ((sum(lambda)^2) + sum(theta))

legacy_model <- 'LegacyPigment =~ 1*a + Lm + C'
legacy_fit <- lavaan::cfa(
  legacy_model,
  data = df2[cfa_rows, c("a", "Lm", "C")],
  estimator = "MLR", std.lv = FALSE, missing = "fiml"
)
legacy_score <- as.numeric(lavaan::lavPredict(legacy_fit, type = "lv")[, "LegacyPigment"])
legacy_cor <- suppressWarnings(cor(score_qc, legacy_score, use = "complete.obs"))

write_csv(loading_v2, file.path(cmp_dir, "pigment_cfa_v2_standardized_loadings.csv"))
write_csv(
  tibble::tibble(metric = names(fit_measures), value = as.numeric(fit_measures)),
  file.path(cmp_dir, "pigment_cfa_v2_fit_measures.csv")
)
write_csv(
  tibble::tibble(
    n_total = nrow(df2),
    n_qc_eligible_complete = length(cfa_rows),
    omega_composite = omega_cfa,
    correlation_with_legacy_score = legacy_cor,
    extraction_version = if ("extraction_version" %in% names(df2))
      paste(sort(unique(na.omit(df2$extraction_version))), collapse = ";") else NA_character_,
    primary_rgb_columns = paste(rgb_cols, collapse = ",")
  ),
  file.path(cmp_dir, "pigment_cfa_v2_summary.csv")
)
write_csv(
  df2[, intersect(c("sample_id", "observation_id", "cfa_qc_eligible", "Pigment",
                    "a", "Lm", "C", "Sm", "qc_status", "qc_flags"), names(df2))],
  file.path(cmp_dir, "pigment_cfa_v2_scores_and_qc.csv")
)

if (is.finite(fit_measures[["cfi.scaled"]]) && fit_measures[["cfi.scaled"]] < 0.90) {
  warning("Pigmentation CFA CFI < 0.90; inspect exported diagnostics before interpretation.")
}
if (is.finite(fit_measures[["rmsea.scaled"]]) && fit_measures[["rmsea.scaled"]] > 0.10) {
  warning("Pigmentation CFA RMSEA > 0.10; inspect exported diagnostics before interpretation.")
}

cat(
  "✓ Upgraded Pigment CFA computed. n =", length(cfa_rows),
  "; omega =", round(omega_cfa, 3),
  "; legacy r =", round(legacy_cor, 3), "\n"
)

colour_pc <- stats::prcomp(cfa_data, center = FALSE, scale. = FALSE)
colour_pc1 <- colour_pc$x[, 1]
if (stats::cor(colour_pc1, score_qc, use = "complete.obs") < 0) colour_pc1 <- -colour_pc1
colour_index_sensitivity <- tibble::tibble(
  comparison = c("CFA_vs_colour_PC1", "CFA_vs_legacy_CFA"),
  correlation = c(
    stats::cor(score_qc, colour_pc1, use = "complete.obs"),
    legacy_cor
  )
)
write_csv(colour_index_sensitivity,
          file.path(cmp_dir, "colour_index_sensitivity.csv"))
```

## Exports

```{r export-loadings}
if (!is.null(temp_pc_obj)) {
  write_csv(
    as.data.frame(temp_pc_obj$rotation) %>% tibble::rownames_to_column("variable"),
    file.path(cmp_dir, "temperature_pca_loadings.csv")
  )
}
if (!is.null(precip_pc_obj)) {
  write_csv(
    as.data.frame(precip_pc_obj$rotation) %>% tibble::rownames_to_column("variable"),
    file.path(cmp_dir, "precip_pca_loadings.csv")
  )
}
if (!is.null(soil_phys_pc_obj)) {
  write_csv(
    as.data.frame(soil_phys_pc_obj$rotation) %>% tibble::rownames_to_column("variable"),
    file.path(cmp_dir, "soil_physical_pca_loadings.csv")
  )
}
if (!is.null(soil_nutrient_pc_obj)) {
  write_csv(
    as.data.frame(soil_nutrient_pc_obj$rotation) %>% tibble::rownames_to_column("variable"),
    file.path(cmp_dir, "soil_nutrient_pca_loadings.csv")
  )
}
if (!is.null(topo_pc_obj)) {
  write_csv(
    as.data.frame(topo_pc_obj$rotation) %>% tibble::rownames_to_column("variable"),
    file.path(cmp_dir, "topography_pca_loadings.csv")
  )
}
```

## Predictors

```{r rsds-pigment-bee}
if (!"RSDS" %in% names(df2)) {
  stop(
    "Required RSDS column is absent after public-raster extraction. ",
    "Check the RSDS acquisition step and public_files names.",
    call. = FALSE
  )
}
df2$RSDS <- df2[["RSDS"]]

stopifnot("Pigment" %in% names(df2), any(is.finite(df2$Pigment)))

bee_cols <- intersect(
  c("ardens", "beaticola", "consobrinus", "diversus", "honshuensis"),
  names(df2)
)
stopifnot(length(bee_cols) > 0)

bee_values <- as.data.frame(lapply(df2[, bee_cols, drop = FALSE], function(x) pmin(pmax(as.numeric(x), 0), 1)))

df2$Bombus_union_score <- apply(bee_values, 1, function(x) {
  x <- x[is.finite(x)]
  if (!length(x)) NA_real_ else 1 - prod(1 - x)
})

df2$Bombus_max  <- apply(
  bee_values, 1,
  function(x) if (all(!is.finite(x))) NA_real_ else max(x, na.rm = TRUE)
)
df2$Bombus_mean <- rowMeans(bee_values, na.rm = TRUE)

df2$Bombus_primary <- df2$Bombus_union_score

bee_index_definition <- tibble::tribble(
  ~index, ~formula, ~interpretation, ~probability_claim,
  "Bombus_union_score", "1 - product(1 - s_i)",
  "bounded union-like composite of five cloglog suitability scores", FALSE,
  "Bombus_max", "max(s_i)",
  "conservative any-species suitability proxy dominated by the best-supported species", FALSE,
  "Bombus_mean", "mean(s_i)",
  "average focal-species suitability", FALSE
)
write_csv(bee_index_definition, file.path(cmp_dir, "bombus_index_definition.csv"))

```

## Model data

```{r build-model-data}
need_vars <- c(
  "Pigment",
  "topo_PC1",
  "Temperature_PC1",
  "precip_PC1",
  "TemperatureSeasonality",
  "PrecipSeasonality",
  "soil_nutrient_PC1",
  "soil_phys_PC1",
  "soil_pH",
  "RSDS",
  "Bombus_primary", "Bombus_max", "Bombus_mean",
  "longitude", "latitude"
)

dat0 <- df2 %>%
  dplyr::select(any_of(c("sample_id", "observation_id", "col_hex")),
                all_of(need_vars), any_of("date"), any_of("elevation")) %>%
  dplyr::filter(complete.cases(across(all_of(need_vars)))) %>%
  dplyr::mutate(
    y         = Pigment,
    z_RSDS    = z(RSDS),
    z_soil_pH = z(soil_pH),
    z_Bee     = z(Bombus_primary),
    z_Bee_max = z(Bombus_max),
    z_Bee_mean = z(Bombus_mean)
  )

stopifnot(nrow(dat0) >= 50)
```

## Diagnostics

```{r predictor-audit, fig.width=8, fig.height=5}
all_candidate_terms <- c(
  "Temperature_PC1",
  "precip_PC1",
  "TemperatureSeasonality",
  "PrecipSeasonality",
  "topo_PC1",
  "soil_phys_PC1",
  "soil_nutrient_PC1",
  "z_soil_pH",
  "z_RSDS",
  "z_Bee"
)
stopifnot(all(all_candidate_terms %in% names(dat0)))

compute_vif <- function(data, terms) {
  purrr::map_dfr(terms, function(term) {
    others <- setdiff(terms, term)
    if (!length(others)) return(tibble::tibble(term = term, VIF = 1))
    fit <- stats::lm(stats::reformulate(others, response = term), data = data)
    r2 <- summary(fit)$r.squared
    tibble::tibble(term = term, VIF = 1 / pmax(1 - r2, 1e-8))
  })
}

filter_vif <- function(data, terms, threshold = 5) {
  kept <- terms
  history <- list()
  removed <- character()
  step <- 0L
  repeat {
    step <- step + 1L
    tab <- compute_vif(data, kept) |> dplyr::arrange(dplyr::desc(VIF), term)
    tab$step <- step
    tab$removed_this_step <- NA_character_
    if (max(tab$VIF, na.rm = TRUE) < threshold) {
      history[[step]] <- tab
      break
    }
    drop_term <- tab$term[1]
    tab$removed_this_step <- drop_term
    history[[step]] <- tab
    removed <- c(removed, drop_term)
    kept <- setdiff(kept, drop_term)
    if (length(kept) < 2L) stop("VIF filtering left fewer than two predictors.", call. = FALSE)
  }
  list(kept = kept, removed = removed, history = dplyr::bind_rows(history))
}

vif_selection <- filter_vif(dat0, all_candidate_terms, threshold = 5)
selected_joint_terms <- vif_selection$kept
selected_abiotic_terms <- setdiff(selected_joint_terms, "z_Bee")
if (!("z_Bee" %in% selected_joint_terms)) {
  stop("Joint VIF filtering removed z_Bee; an independent Bombus extension is not estimable with VIF < 5.", call. = FALSE)
}
stopifnot(length(selected_abiotic_terms) >= 1)

vif_final <- compute_vif(dat0, selected_joint_terms) |> dplyr::arrange(dplyr::desc(VIF))
audit_cor <- stats::cor(dat0[, selected_joint_terms, drop = FALSE], use = "pairwise.complete.obs")
bombus_cor <- tibble::tibble(
  environmental_term = selected_abiotic_terms,
  r_with_Bombus = audit_cor[selected_abiotic_terms, "z_Bee"]
)

write_csv(vif_selection$history, file.path(cmp_dir, "VIF_filter_history.csv"))
write_csv(vif_final, file.path(cmp_dir, "VIF_final_environment_plus_Bombus.csv"))
write_csv(tibble::tibble(
  term = all_candidate_terms,
  retained = all_candidate_terms %in% selected_joint_terms,
  removal_order = match(all_candidate_terms, vif_selection$removed)
), file.path(cmp_dir, "VIF_term_selection.csv"))
utils::write.csv(audit_cor, file.path(cmp_dir, "final_predictor_correlation_matrix.csv"), row.names = TRUE)
write_csv(bombus_cor, file.path(cmp_dir, "Bombus_environment_correlations.csv"))

print(vif_final)
print(tibble::tibble(retained_term = selected_joint_terms))

p_vif <- ggplot(vif_final, aes(VIF, forcats::fct_reorder(term, VIF))) +
  geom_vline(xintercept = 5, linetype = 2) +
  geom_point(size = 2.5) +
  labs(x = "Final VIF", y = NULL, title = "Joint VIF filter: environment + Bombus") +
  theme_classic(base_size = 11)
print(p_vif)
```

## Models

```{r primary-model-definitions}
environment_terms <- selected_abiotic_terms
primary_terms <- list(
  M0_spatial = character(0),
  M1_abiotic_spatial = environment_terms,
  M2_abiotic_Bombus_spatial = c(environment_terms, "z_Bee")
)

write_csv(
  tibble::tibble(
    model = names(primary_terms),
    role = c("spatial reference", "confirmatory abiotic baseline", "exploratory biotic extension"),
    terms = vapply(primary_terms, function(x) paste(x, collapse = " + "), character(1))
  ),
  file.path(cmp_dir, "primary_model_definitions.csv")
)
```

## SPDE model

A 20-km maximum inner mesh edge was used for the final analysis. In a preliminary numerical-resolution check, the 10- and 20-km meshes had effectively identical WAIC, whereas the 40-km mesh performed worse. The 20-km mesh used approximately one quarter of the vertices required by the 10-km mesh and was therefore retained as the parsimonious numerical resolution. Mesh edge length is a computational setting and is not interpreted as an ecological dispersal distance.

```{r spde-model}
sf_cmp <- sf::st_as_sf(dat0, coords = c("longitude", "latitude"), crs = 4326)
sf_cmp_m <- sf::st_transform(sf_cmp, 3857)
coords_cmp <- sf::st_coordinates(sf_cmp_m)

mesh_cmp <- INLA::inla.mesh.2d(
  loc = coords_cmp,
  max.edge = 1000 * c(20, 100),
  cutoff = 1000 * 5
)

spde_cmp <- INLA::inla.spde2.matern(mesh_cmp, alpha = 2)
A_cmp <- INLA::inla.spde.make.A(mesh_cmp, loc = coords_cmp)

selected_bundle <- list(
  mesh = mesh_cmp,
  spde = spde_cmp,
  A = A_cmp
)

fit_spde <- function(dat, terms, model_name, bundle = selected_bundle) {
  X <- if (length(terms)) {
    dat[, terms, drop = FALSE]
  } else {
    data.frame(row.names = seq_len(nrow(dat)))
  }

  X$Intercept <- 1
  X <- X[, c("Intercept", terms), drop = FALSE]

  stk <- INLA::inla.stack(
    data = list(y = dat$y),
    A = list(bundle$A, 1),
    effects = list(
      s = seq_len(bundle$spde$n.spde),
      X = X
    ),
    tag = "est"
  )

  rhs <- c("-1", names(X), "f(s, model = bundle$spde)")
  formula <- stats::as.formula(
    paste("y ~", paste(rhs, collapse = " + "))
  )

  fit_capture <- capture_warnings(
    INLA::inla(
      formula,
      data = INLA::inla.stack.data(stk),
      family = "gaussian",
      control.predictor = list(
        A = INLA::inla.stack.A(stk),
        compute = TRUE
      ),
      control.compute = list(
        config = TRUE,
        dic = TRUE,
        waic = TRUE,
        cpo = TRUE
      )
    )
  )

  fit <- fit_capture$value
  idx <- INLA::inla.stack.index(stk, "est")$data

  stats <- tibble::tibble(
    model = model_name,
    DIC = fit$dic$dic,
    WAIC = fit$waic$waic,
    pWAIC = fit$waic$p.eff,
    mean_neglogCPO = mean(-log(fit$cpo$cpo[idx]), na.rm = TRUE),
    n_CPO_NA = sum(!is.finite(fit$cpo$cpo[idx]))
  )

  list(
    name = model_name,
    terms = terms,
    formula = formula,
    fit = fit,
    stack = stk,
    index = idx,
    stats = stats,
    warnings = fit_capture$warnings,
    bundle = bundle
  )
}

primary_fits <- purrr::imap(
  primary_terms,
  ~fit_spde(dat0, .x, .y)
)

m1 <- primary_fits[["M1_abiotic_spatial"]]
m2 <- primary_fits[["M2_abiotic_Bombus_spatial"]]
delta_m2_vs_m1 <- m2$stats$WAIC - m1$stats$WAIC

primary_comparison <- dplyr::bind_rows(
  purrr::map(primary_fits, "stats")
) |>
  dplyr::arrange(WAIC) |>
  dplyr::mutate(
    delta_WAIC = WAIC - min(WAIC, na.rm = TRUE)
  )

write_csv(
  primary_comparison,
  file.path(cmp_dir, "M0_M1_M2_comparison.csv")
)

model_warnings <- purrr::imap_dfr(
  primary_fits,
  function(x, nm) {
    if (!length(x$warnings)) return(tibble::tibble())
    tibble::tibble(model = nm, warning = x$warnings)
  }
)

print(primary_comparison)
if (nrow(model_warnings)) {
  print(model_warnings)
} else {
  cat("No INLA warnings.\n")
}

write_csv(
  model_warnings,
  file.path(cmp_dir, "model_warnings.csv")
)
```

## Selected model

```{r select-best-model}
best_model_name <- primary_comparison$model[which.min(primary_comparison$WAIC)]
best_model <- primary_fits[[best_model_name]]

model_selection <- primary_comparison |>
  dplyr::mutate(
    selected = model == best_model_name,
    equivalent_support = delta_WAIC < 2
  )
write_csv(model_selection, file.path(best_dir, "selected_model.csv"))
print(model_selection)
```

## Best-model results

```{r best-model-results}
fit_final <- best_model$fit
stk_final <- best_model$stack
IDX <- INLA::inla.stack.index(stk_final, "est")$data

# Fixed effects
fixef_final <- fit_final$summary.fixed |>
  as.data.frame() |>
  tibble::rownames_to_column("term")

write_csv(fixef_final, file.path(best_dir, "best_model_fixed_effects.csv"))
print(fixef_final)

# Observation-level spatial field and fixed linear predictor
spatial_nodes <- fit_final$summary.random$s$mean
s_obs <- as.numeric(A_cmp %*% spatial_nodes)

eta <- fit_final$summary.linear.predictor$mean[IDX]
fixed_part <- eta - s_obs

# Gaussian observation variance: posterior median precision
precision_row <- grep(
  "Precision.*Gaussian",
  rownames(fit_final$summary.hyperpar),
  value = TRUE
)[1]

if (is.na(precision_row) || !nzchar(precision_row)) {
  stop("Gaussian observation precision was not found in fit_final$summary.hyperpar.")
}

prec_med <- fit_final$summary.hyperpar[precision_row, "0.5quant"]
resid_var <- 1 / prec_med

# Standard descriptive variance decomposition
v_fixed   <- stats::var(fixed_part, na.rm = TRUE)
v_spatial <- stats::var(s_obs, na.rm = TRUE)
v_resid   <- resid_var
v_total   <- v_fixed + v_spatial + v_resid

var_decomp <- tibble::tibble(
  component = c("Fixed effects", "Spatial field", "Residual"),
  variance = c(v_fixed, v_spatial, v_resid),
  proportion = c(v_fixed, v_spatial, v_resid) / v_total,
  percent = 100 * proportion,
  selected_model = best_model_name
)

write_csv(
  var_decomp,
  file.path(best_dir, "best_model_variance_decomposition.csv")
)
print(var_decomp)

p_var_decomp <- ggplot(
  var_decomp,
  aes(x = "Selected model", y = proportion, fill = component)
) +
  geom_col(width = 0.62) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    x = NULL,
    y = "Proportion of component variance",
    fill = NULL,
    title = paste("Variance decomposition:", best_model_name),
    subtitle = "Fixed linear predictor, SPDE field, and Gaussian observation variance"
  ) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "bottom"
  )

print(p_var_decomp)
ggsave(
  file.path(best_plotdir, "best_model_variance_decomposition.png"),
  p_var_decomp,
  width = 7.2,
  height = 4.8,
  dpi = 300
)

# Mesh specification retained for reproducibility
mesh_specification <- tibble::tibble(
  inner_max_edge_km = 20,
  outer_max_edge_km = 100,
  cutoff_km = 5,
  mesh_vertices = mesh_cmp$n,
  interpretation = "Numerical resolution only; not an ecological distance."
)

write_csv(
  mesh_specification,
  file.path(best_dir, "best_model_mesh_specification.csv")
)
print(mesh_specification)

# Fitted values and residuals
y_obs   <- INLA::inla.stack.data(stk_final)$y[IDX]
yhat    <- fit_final$summary.fitted.values$mean[IDX]
yhat_sd <- fit_final$summary.fitted.values$sd[IDX]

resid_raw  <- y_obs - yhat
resid_zfit <- resid_raw / pmax(as.numeric(yhat_sd), 1e-6)

dat_used <- dat0
dat_used$yhat       <- yhat
dat_used$yhat_sd    <- yhat_sd
dat_used$resid_raw  <- resid_raw
dat_used$resid_zfit <- resid_zfit

write_csv(dat_used, file.path(best_dir, "best_model_residuals.csv"))

response_scale_check <- tibble::tibble(
  statistic = c(
    "n",
    "mean response",
    "SD response",
    "variance response",
    "fixed + spatial + residual variance"
  ),
  value = c(
    length(y_obs),
    mean(y_obs, na.rm = TRUE),
    stats::sd(y_obs, na.rm = TRUE),
    stats::var(y_obs, na.rm = TRUE),
    v_total
  )
)

write_csv(
  response_scale_check,
  file.path(best_dir, "best_model_response_scale_check.csv")
)
print(response_scale_check)

# Fixed-effect figure
coef_plot_tbl <- fixef_final |>
  dplyr::filter(term != "Intercept") |>
  dplyr::mutate(
    CI_excludes_zero = (`0.025quant` > 0) | (`0.975quant` < 0),
    term = forcats::fct_reorder(term, mean)
  )

p_fixef <- ggplot(coef_plot_tbl, aes(mean, term)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey55") +
  geom_errorbarh(
    aes(xmin = `0.025quant`, xmax = `0.975quant`),
    height = 0.12
  ) +
  geom_point(
    aes(fill = CI_excludes_zero),
    shape = 21,
    size = 3,
    stroke = 0.7
  ) +
  scale_fill_manual(
    values = c(`FALSE` = "white", `TRUE` = "black"),
    guide = "none"
  ) +
  labs(
    x = "Posterior mean coefficient (95% CrI)",
    y = NULL,
    title = paste("Selected model:", best_model_name)
  ) +
  theme_classic(base_size = 11)

print(p_fixef)
ggsave(
  file.path(best_plotdir, "selected_model_fixed_effects.png"),
  p_fixef,
  width = 8,
  height = 5.5,
  dpi = 300
)
```

## Natural anomalies

```{r natural-anomalies}
y_obs <- INLA::inla.stack.data(stk_final)$y[IDX]
yhat <- fit_final$summary.fitted.values$mean[IDX]
yhat_sd <- fit_final$summary.fitted.values$sd[IDX]
pit <- fit_final$cpo$pit[IDX]
pit_clamped <- pmin(pmax(as.numeric(pit), 1e-6), 1 - 1e-6)

anomaly <- stats::qnorm(pit_clamped)
dat_used <- dat0
dat_used$natural_expected <- yhat
dat_used$natural_expected_sd <- yhat_sd
dat_used$natural_anomaly <- anomaly
dat_used$upper_anomaly_95 <- anomaly > stats::qnorm(0.95)
dat_used$lower_anomaly_05 <- anomaly < stats::qnorm(0.05)

write_csv(dat_used, file.path(best_dir, "natural_model_anomalies.csv"))
write_csv(
  tibble::tibble(
    selected_model = best_model_name,
    n = nrow(dat_used),
    n_upper_95 = sum(dat_used$upper_anomaly_95, na.rm = TRUE),
    n_lower_05 = sum(dat_used$lower_anomaly_05, na.rm = TRUE),
    PIT_missing = sum(!is.finite(pit))
  ),
  file.path(best_dir, "anomaly_summary.csv")
)
```

## Upper-tail analysis

```{r anomaly-qgam}
set.seed(42)

if ("date" %in% names(dat_used)) {
  date_parsed <- suppressWarnings(as.Date(dat_used$date, tryFormats = c(
    "%Y-%m-%d", "%Y/%m/%d", "%Y.%m.%d", "%Y%m%d",
    "%Y-%m-%d %H:%M", "%Y/%m/%d %H:%M"
  )))
  dat_used$DOY <- as.integer(strftime(date_parsed, "%j"))
} else dat_used$DOY <- NA_integer_

pts_human <- terra::vect(dat_used, geom = c("longitude", "latitude"), crs = "EPSG:4326")
r_human <- terra::rast(worldpop_file)
if (!terra::same.crs(pts_human, r_human)) pts_human <- terra::project(pts_human, terra::crs(r_human))
dat_used$population_count <- terra::extract(r_human, pts_human, ID = FALSE)[, 1]
dat_used$human_exposure <- log1p(dat_used$population_count)

q_required <- c("natural_anomaly", "longitude", "latitude", "human_exposure", "elevation")
if (sum(is.finite(dat_used$DOY)) >= 100 && length(unique(na.omit(dat_used$DOY))) >= 8) q_required <- c(q_required, "DOY")
qdat <- dat_used[stats::complete.cases(dat_used[, q_required]), , drop = FALSE]

q_formula <- if ("DOY" %in% q_required) {
  natural_anomaly ~ s(longitude, latitude, k = 60) + s(elevation, k = 10) +
    s(human_exposure, k = 10) + s(DOY, k = 12)
} else {
  natural_anomaly ~ s(longitude, latitude, k = 60) + s(elevation, k = 10) +
    s(human_exposure, k = 10)
}

taus <- c(0.50, 0.75, 0.90, 0.95)
qgam_capture <- lapply(taus, function(tau) capture_warnings(
  qgam::qgam(q_formula, data = qdat, qu = tau)
))
qgam_models <- stats::setNames(lapply(qgam_capture, `[[`, "value"), paste0("q", taus))
qgam_warnings <- purrr::map2_dfr(qgam_capture, taus, function(x, tau) {
  if (!length(x$warnings)) return(tibble::tibble())
  tibble::tibble(tau = tau, warning = x$warnings)
})
write_csv(qgam_warnings, file.path(gam_dir, "qGAM_warnings.csv"))
if (nrow(qgam_warnings)) print(qgam_warnings) else cat("No qGAM warnings.\n")

qgam_summary <- purrr::imap_dfr(qgam_models, function(mod, nm) {
  sm <- summary(mod)$s.table
  tibble::as_tibble(sm, rownames = "smooth") |>
    dplyr::mutate(model = nm, tau = as.numeric(sub("q", "", nm)), .before = 1)
})
write_csv(qgam_summary, file.path(gam_dir, "upper_tail_qGAM_smooth_tests.csv"))
writeLines(paste(deparse(q_formula), collapse = " "), file.path(gam_dir, "qGAM_formula.txt"))
```

### qGAM effect plots

```{r qgam-effect-plots}
make_qgam_newdata <- function(data, varname, n = 200) {
  xseq <- seq(
    min(data[[varname]], na.rm = TRUE),
    max(data[[varname]], na.rm = TRUE),
    length.out = n
  )

  nd <- data[rep(1, n), , drop = FALSE]
  nd[[varname]] <- xseq

  hold_vars <- intersect(
    c("longitude", "latitude", "elevation", "human_exposure", "DOY"),
    names(nd)
  )
  hold_vars <- setdiff(hold_vars, varname)

  for (v in hold_vars) {
    nd[[v]] <- median(data[[v]], na.rm = TRUE)
  }

  list(x = xseq, newdata = nd)
}

get_qgam_curve <- function(mod, data, varname, tau_label, n = 200) {
  nd <- make_qgam_newdata(data, varname, n)
  pr <- predict(mod, newdata = nd$newdata, se.fit = TRUE, type = "response")

  tibble::tibble(
    variable = varname,
    x = nd$x,
    fit = as.numeric(pr$fit),
    se = as.numeric(pr$se.fit),
    lower = fit - 1.96 * se,
    upper = fit + 1.96 * se,
    tau = tau_label
  )
}

q95_name <- names(qgam_models)[which.min(abs(taus - 0.95))]
q95_model <- qgam_models[[q95_name]]

qgam_plot_vars <- intersect(
  c("human_exposure", "DOY", "elevation"),
  q_required
)

q95_curves <- purrr::map_dfr(
  qgam_plot_vars,
  ~get_qgam_curve(q95_model, qdat, .x, "0.95")
)

q95_plots <- purrr::map(
  qgam_plot_vars,
  function(v) {
    dfp <- dplyr::filter(q95_curves, variable == v)
    p <- ggplot(dfp, aes(x, fit)) +
      geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.20) +
      geom_line(linewidth = 0.9) +
      geom_hline(yintercept = 0, linetype = 2, colour = "grey55") +
      labs(
        x = dplyr::recode(
          v,
          human_exposure = "Human exposure (log population count)",
          DOY = "Day of year",
          elevation = "Elevation"
        ),
        y = "Predicted 0.95 anomaly quantile",
        title = paste("Upper-tail effect:", v)
      ) +
      theme_classic(base_size = 11)

    print(p)
    ggsave(
      file.path(gam_dir, paste0("qGAM_q095_", v, ".png")),
      p, width = 6.2, height = 4.5, dpi = 300
    )
    p
  }
)
names(q95_plots) <- qgam_plot_vars

multi_tau_curves <- purrr::imap_dfr(
  qgam_models,
  function(mod, nm) {
    purrr::map_dfr(
      qgam_plot_vars,
      ~get_qgam_curve(
        mod,
        qdat,
        .x,
        as.numeric(sub("q", "", nm))
      )
    )
  }
)

multi_tau_plots <- purrr::map(
  qgam_plot_vars,
  function(v) {
    dfp <- dplyr::filter(multi_tau_curves, variable == v)
    p <- ggplot(dfp, aes(x, fit, group = tau, linetype = factor(tau))) +
      geom_line(linewidth = 0.9) +
      geom_hline(yintercept = 0, linetype = 2, colour = "grey65") +
      labs(
        x = dplyr::recode(
          v,
          human_exposure = "Human exposure (log population count)",
          DOY = "Day of year",
          elevation = "Elevation"
        ),
        y = "Predicted anomaly quantile",
        linetype = expression(tau),
        title = paste("Quantile-specific effect:", v)
      ) +
      theme_classic(base_size = 11) +
      theme(legend.position = "bottom")

    print(p)
    ggsave(
      file.path(gam_dir, paste0("qGAM_multitau_", v, ".png")),
      p, width = 6.5, height = 4.6, dpi = 300
    )
    p
  }
)
names(multi_tau_plots) <- qgam_plot_vars

write_csv(q95_curves, file.path(gam_dir, "qGAM_q095_effect_curves.csv"))
write_csv(multi_tau_curves, file.path(gam_dir, "qGAM_multitau_effect_curves.csv"))
```

## Candidates

```{r candidate-ranking}
rank01 <- function(x, decreasing = FALSE) {
  r <- rank(if (decreasing) -x else x, na.last = "keep", ties.method = "average")
  (r - 1) / pmax(sum(is.finite(r)) - 1, 1)
}

candidates <- dat_used |>
  dplyr::mutate(
    anomaly_percentile = rank01(natural_anomaly),
    human_percentile = rank01(human_exposure),
    early_percentile = if ("DOY" %in% names(dat_used)) rank01(DOY, decreasing = TRUE) else NA_real_,
    horticultural_candidate_score =
      0.60 * anomaly_percentile +
      0.25 * human_percentile +
      0.15 * dplyr::coalesce(early_percentile, 0.5)
  ) |>
  dplyr::arrange(dplyr::desc(horticultural_candidate_score))

write_csv(candidates, file.path(gam_dir, "horticultural_candidate_ranking.csv"))
write_csv(head(candidates, 100), file.path(gam_dir, "top100_field_validation_candidates.csv"))
```

## Sensitivity

```{r minimal-sensitivity}
bee_sensitivity_terms <- c(z_Bee_max = "z_Bee_max", z_Bee_mean = "z_Bee_mean")
bee_sensitivity <- purrr::imap_dfr(bee_sensitivity_terms, function(term, label) {
  obj <- fit_spde(dat0, c(environment_terms, term), paste0("sensitivity_", label), selected_bundle)
  row <- obj$fit$summary.fixed[term, , drop = FALSE]
  tibble::tibble(
    index = label,
    posterior_mean = row[, "mean"],
    lower_95 = row[, "0.025quant"],
    upper_95 = row[, "0.975quant"],
    WAIC = obj$stats$WAIC,
    delta_WAIC_vs_M1 = obj$stats$WAIC - m1$stats$WAIC
  )
})
write_csv(bee_sensitivity, file.path(top_dir, "Bombus_index_sensitivity.csv"))
```

## Submission files

```{r submission-materials}
term_labels <- c(
  Temperature_PC1 = "Thermal growing conditions",
  precip_PC1 = "Moisture amount / dryness",
  TemperatureSeasonality = "Temperature seasonality",
  PrecipSeasonality = "Precipitation seasonality",
  topo_PC1 = "Topographic complexity",
  soil_phys_PC1 = "Soil physical properties",
  soil_nutrient_PC1 = "Soil nutrient status",
  z_soil_pH = "Soil pH",
  z_RSDS = "Solar radiation",
  z_Bee = "Bombus-associated suitability"
)

fig1_dat <- dat0 |>
  dplyr::mutate(display_colour = dplyr::coalesce(as.character(col_hex), "#808080"))

p1a <- ggplot(fig1_dat, aes(longitude, latitude)) +
  geom_point(aes(colour = display_colour), size = 1.1, alpha = 0.75) +
  scale_colour_identity() +
  coord_quickmap() +
  labs(x = "Longitude", y = "Latitude", tag = "a",
       title = "Nationwide georeferenced floral colour") +
  theme_classic(base_size = 10)

p1b <- ggplot(fig1_dat, aes(y)) +
  geom_histogram(bins = 40, boundary = 0) +
  labs(x = "Latent floral-colour index", y = "Observations", tag = "b",
       title = "Continuous pigmentation variation") +
  theme_classic(base_size = 10)

cfa_plot_tbl <- loading_v2 |>
  dplyr::mutate(
    indicator = dplyr::recode(rhs,
      a_z = "Red–magenta (a*)",
      Lm_z = "Inverted lightness",
      C_z = "Chroma",
      Sm_z = "Saturation"
    ),
    indicator = forcats::fct_reorder(indicator, est.std)
  )
p1c <- ggplot(cfa_plot_tbl, aes(est.std, indicator)) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_point(size = 2.5) +
  labs(x = "Standardized loading", y = NULL, tag = "c",
       title = "Measurement model") +
  theme_classic(base_size = 10)

if (requireNamespace("patchwork", quietly = TRUE)) {
  figure1 <- (p1a | p1b) / p1c +
    patchwork::plot_layout(heights = c(1.15, 0.75), guides = "collect") +
    patchwork::plot_annotation(title = "Figure 1. Georeferenced phenotyping and continuous floral-colour variation") &
    theme(plot.title = element_text(face = "bold", size = 11),
          plot.tag = element_text(face = "bold"),
          legend.position = "bottom")
  save_pub_plot(figure1, "Figure_1_phenotyping_and_colour", 178/25.4, 175/25.4)
  print(figure1)
} else {
  save_pub_plot(p1a, "Figure_1a_colour_map", 90/25.4, 90/25.4)
  save_pub_plot(p1b, "Figure_1b_pigment_distribution", 90/25.4, 90/25.4)
  save_pub_plot(p1c, "Figure_1c_CFA_loadings", 90/25.4, 70/25.4)
}

coef_fig_tbl <- best_model$fit$summary.fixed |>
  as.data.frame() |>
  tibble::rownames_to_column("term") |>
  dplyr::filter(term != "Intercept") |>
  dplyr::mutate(
    label = dplyr::recode(term, !!!term_labels),
    label = forcats::fct_reorder(label, mean),
    supported = (`0.025quant` > 0) | (`0.975quant` < 0)
  )
p2a <- ggplot(coef_fig_tbl, aes(mean, label)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey55") +
  geom_errorbarh(aes(xmin = `0.025quant`, xmax = `0.975quant`), height = 0.12) +
  geom_point(aes(fill = supported), shape = 21, size = 2.8, stroke = 0.7) +
  scale_fill_manual(values = c(`FALSE` = "white", `TRUE` = "black"), guide = "none") +
  labs(x = "Posterior coefficient (95% CrI)", y = NULL, tag = "a",
       title = paste("Fixed effects:", best_model_name)) +
  theme_classic(base_size = 10)

mcomp <- primary_comparison |>
  dplyr::mutate(
    label = dplyr::recode(model,
      M0_spatial = "M0: spatial",
      M1_abiotic_spatial = "M1: abiotic + spatial",
      M2_abiotic_Bombus_spatial = "M2: abiotic + Bombus + spatial"
    ),
    label = forcats::fct_reorder(label, delta_WAIC, .desc = TRUE)
  )
p2b <- ggplot(mcomp, aes(delta_WAIC, label)) +
  geom_col() +
  labs(x = expression(Delta*"WAIC"), y = NULL, tag = "b",
       title = "Prespecified model comparison") +
  theme_classic(base_size = 10)

if (requireNamespace("patchwork", quietly = TRUE)) {
  figure2 <- (p2a | p2b) / p_var_decomp +
    patchwork::plot_layout(heights = c(1.15, 0.70), widths = c(1.55, 1), guides = "collect") +
    patchwork::plot_annotation(title = "Figure 2. Selected ecological–spatial model") &
    theme(plot.title = element_text(face = "bold", size = 11),
          plot.tag = element_text(face = "bold"),
          legend.position = "bottom")
  save_pub_plot(figure2, "Figure_2_selected_model", 178/25.4, 155/25.4)
  print(figure2)
} else {
  save_pub_plot(p2a, "Figure_2a_abiotic_coefficients", 110/25.4, 100/25.4)
  save_pub_plot(p2b, "Figure_2b_model_comparison", 68/25.4, 100/25.4)
}

bee_main <- m2$fit$summary.fixed["z_Bee", , drop = FALSE]
bee_plot <- dplyr::bind_rows(
  tibble::tibble(
    index = "Union-like primary",
    posterior_mean = bee_main[, "mean"],
    lower_95 = bee_main[, "0.025quant"],
    upper_95 = bee_main[, "0.975quant"],
    delta_WAIC_vs_M1 = m2$stats$WAIC - m1$stats$WAIC
  ),
  bee_sensitivity |>
    dplyr::transmute(
      index = dplyr::recode(index,
        z_Bee_max = "Maximum",
        z_Bee_mean = "Mean"
      ),
      posterior_mean, lower_95, upper_95, delta_WAIC_vs_M1
    )
) |>
  dplyr::mutate(index = forcats::fct_reorder(index, posterior_mean))

p3a <- ggplot(bee_plot, aes(posterior_mean, index)) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_errorbarh(aes(xmin = lower_95, xmax = upper_95), height = 0.15) +
  geom_point(size = 2.6) +
  labs(x = "Bombus coefficient (95% credible interval)", y = NULL, tag = "a",
       title = "Sensitivity to composite definition") +
  theme_classic(base_size = 10)

p3b <- ggplot(bombus_cor,
              aes(r_with_Bombus,
                  forcats::fct_reorder(dplyr::recode(environmental_term, !!!term_labels),
                                      r_with_Bombus))) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_point(size = 2.6) +
  labs(x = "Correlation with primary Bombus index", y = NULL, tag = "b",
       title = "Overlap with measured environment") +
  theme_classic(base_size = 10)

if (requireNamespace("patchwork", quietly = TRUE)) {
  figure3 <- p3a | p3b +
    patchwork::plot_layout(widths = c(1.15, 1), guides = "collect") +
    patchwork::plot_annotation(title = "Figure 3. Exploratory pollinator-associated context") &
    theme(plot.title = element_text(face = "bold", size = 11),
          plot.tag = element_text(face = "bold"),
          legend.position = "bottom")
  save_pub_plot(figure3, "Figure_3_Bombus_context", 178/25.4, 105/25.4)
  print(figure3)
} else {
  save_pub_plot(p3a, "Figure_3a_Bombus_sensitivity", 90/25.4, 95/25.4)
  save_pub_plot(p3b, "Figure_3b_Bombus_environment_overlap", 88/25.4, 95/25.4)
}

p4a <- ggplot(dat_used, aes(longitude, latitude, colour = natural_anomaly)) +
  geom_point(size = 1.25, alpha = 0.8) +
  coord_quickmap() +
  scale_colour_gradient2(midpoint = 0) +
  labs(x = "Longitude", y = "Latitude", colour = "Selected-model\nanomaly",
       tag = "a", title = "Deviation from the selected ecological–spatial model") +
  theme_classic(base_size = 10)

upper_bin <- dat_used |>
  dplyr::filter(is.finite(human_exposure), !is.na(upper_anomaly_95)) |>
  dplyr::mutate(exposure_bin = dplyr::ntile(human_exposure, 5L)) |>
  dplyr::group_by(exposure_bin) |>
  dplyr::summarise(
    human_exposure = stats::median(human_exposure, na.rm = TRUE),
    upper_tail_fraction = mean(upper_anomaly_95, na.rm = TRUE),
    n = dplyr::n(), .groups = "drop"
  )

p4b <- ggplot(upper_bin, aes(human_exposure, upper_tail_fraction)) +
  geom_hline(yintercept = 0.05, linetype = 2) +
  geom_line() + geom_point(size = 2.5) +
  labs(x = "Human exposure, log(population + 1)",
       y = "Fraction above selected-model 95% threshold", tag = "b",
       title = "Upper-tail enrichment across human exposure") +
  theme_classic(base_size = 10)

q_effect <- purrr::imap_dfr(qgam_models, function(mod, nm) {
  xseq <- seq(stats::quantile(qdat$human_exposure, 0.02, na.rm = TRUE),
              stats::quantile(qdat$human_exposure, 0.98, na.rm = TRUE),
              length.out = 120)
  nd <- data.frame(
    longitude = stats::median(qdat$longitude, na.rm = TRUE),
    latitude = stats::median(qdat$latitude, na.rm = TRUE),
    elevation = stats::median(qdat$elevation, na.rm = TRUE),
    human_exposure = xseq
  )
  if ("DOY" %in% names(qdat)) nd$DOY <- stats::median(qdat$DOY, na.rm = TRUE)
  pr <- stats::predict(mod, newdata = nd, se.fit = TRUE)
  tibble::tibble(
    human_exposure = xseq,
    fit = as.numeric(pr$fit),
    lower = as.numeric(pr$fit - 1.96 * pr$se.fit),
    upper = as.numeric(pr$fit + 1.96 * pr$se.fit),
    tau = as.numeric(sub("q", "", nm))
  )
})
p4c <- ggplot(q_effect, aes(human_exposure, fit, group = factor(tau),
                            linetype = factor(tau))) +
  geom_line(linewidth = 0.8) +
  labs(x = "Human exposure, log(population + 1)",
       y = "Predicted anomaly", linetype = expression(tau),
       tag = "c", title = "Conditional quantile responses") +
  theme_classic(base_size = 10)

if (requireNamespace("patchwork", quietly = TRUE)) {
  p4c_final <- if ("human_exposure" %in% names(q95_plots)) {
    q95_plots[["human_exposure"]]
  } else {
    p4c
  }

  figure4 <- (p4a | p4b) / p4c_final +
    patchwork::plot_layout(heights = c(1.1, 0.85), widths = c(1.25, 1), guides = "collect") +
    patchwork::plot_annotation(title = "Figure 4. Upper-tail deviations from the selected ecological–spatial model") &
    theme(plot.title = element_text(face = "bold", size = 11),
          plot.tag = element_text(face = "bold"),
          legend.position = "bottom")
  save_pub_plot(figure4, "Figure_4_natural_anomalies", 178/25.4, 175/25.4)
  print(figure4)
} else {
  save_pub_plot(p4a, "Figure_4a_anomaly_map", 90/25.4, 90/25.4)
  save_pub_plot(p4b, "Figure_4b_anomaly_scatter", 88/25.4, 90/25.4)
  save_pub_plot(p4c, "Figure_4c_quantile_effects", 120/25.4, 80/25.4)
}

table1 <- tibble::tibble(
  domain = c("Thermal growing conditions", "Temperature seasonality",
             "Moisture amount / dryness", "Precipitation seasonality",
             "Topography", "Soil physical", "Soil nutrient", "Soil chemistry",
             "Solar radiation", "Pollinator-associated context"),
  variables = c(
    paste(temp_vars, collapse = ", "),
    "BIO4",
    paste(precip_vars, collapse = ", "),
    "BIO15",
    paste(topo_vars, collapse = ", "),
    paste(soil_phys_vars, collapse = ", "),
    paste(soil_nutrient_vars, collapse = ", "),
    "Soil pH",
    "RSDS",
    "Five species-specific Bombus SDMs; union-like composite"
  ),
  representation = c(
    "PC1", "standardized single predictor", "PC1",
    "standardized single predictor", "PC1", "PC1", "PC1",
    "standardized single predictor", "standardized single predictor",
    "exploratory standardized index"
  )
)
write_csv(table1, file.path(table_dir, "Table_1_predictor_definitions.csv"))
write_csv(primary_comparison, file.path(table_dir, "Table_2_primary_model_comparison.csv"))
write_csv(fixef_final, file.path(table_dir, "Table_3_selected_model_coefficients.csv"))
write_csv(var_decomp, file.path(table_dir, "Table_4_variance_decomposition.csv"))
write_csv(qgam_summary, file.path(table_dir, "Table_5_qGAM_smooth_tests.csv"))
write_csv(bee_plot, file.path(table_dir, "Table_S_Bombus_index_sensitivity.csv"))

sdm_selected_files <- list.files(bombus_eval_dir, pattern = "_selected_model\\\\.csv$",
                                 full.names = TRUE)
if (length(sdm_selected_files)) {
  sdm_selected <- dplyr::bind_rows(lapply(sdm_selected_files, readr::read_csv,
                                          show_col_types = FALSE))
  write_csv(sdm_selected, file.path(table_dir, "Table_S_Bombus_ENMeval_selected_models.csv"))
}

cfa_values <- setNames(as.numeric(fit_measures), names(fit_measures))
reviewer_audit <- tibble::tribble(
  ~component, ~check, ~value, ~status,
  "Colour measurement", "CFA convergence", as.character(lavaan::inspect(fit_pig, "converged")),
  ifelse(isTRUE(lavaan::inspect(fit_pig, "converged")), "pass", "fail"),
  "Colour measurement", "Composite omega", sprintf("%.3f", omega_cfa),
  ifelse(is.finite(omega_cfa) && omega_cfa >= 0.70, "pass", "inspect"),
  "Colour measurement", "Correlation with colour PC1",
  sprintf("%.3f", colour_index_sensitivity$correlation[1]),
  ifelse(colour_index_sensitivity$correlation[1] >= 0.90, "pass", "inspect"),
  "Predictors", "Maximum absolute pairwise correlation",
  sprintf("%.3f", max(abs(audit_cor[upper.tri(audit_cor)]), na.rm = TRUE)),
  ifelse(max(abs(audit_cor[upper.tri(audit_cor)]), na.rm = TRUE) < 0.70, "pass", "inspect"),
  "Predictors", "Maximum VIF including Bombus",
  sprintf("%.3f", max(vif_final$VIF, na.rm = TRUE)),
  ifelse(max(vif_final$VIF, na.rm = TRUE) < 5, "pass", "inspect"),
  "Spatial models", "CPO failures in selected baseline",
  as.character(sum(!is.finite(fit_final$cpo$cpo[IDX]))),
  ifelse(sum(!is.finite(fit_final$cpo$cpo[IDX])) == 0, "pass", "inspect"),
  "Natural anomaly", "PIT values available",
  sprintf("%d/%d", sum(is.finite(pit)), length(pit)),
  ifelse(mean(is.finite(pit)) >= 0.99, "pass", "inspect"),
  "Bombus", "M2 improves exploratory fit", as.character(delta_m2_vs_m1 < 0),
  "descriptive_only",
  "Anthropogenic exploration", "qGAM complete observations",
  as.character(nrow(qdat)),
  ifelse(nrow(qdat) >= 500, "pass", "inspect")
)
write_csv(reviewer_audit, file.path(repro_dir, "reviewer_audit.csv"))

fmt <- function(x, digits = 2) formatC(x, digits = digits, format = "f")
m0_waic <- primary_fits[["M0_spatial"]]$stats$WAIC
m1_waic <- m1$stats$WAIC
m2_waic <- m2$stats$WAIC
supported_terms <- coef_fig_tbl |>
  dplyr::filter(supported) |>
  dplyr::pull(label)

results_lines <- c(
  paste0("A total of ", nrow(dat0),
         " QC-eligible georeferenced observations entered the spatial analysis."),
  paste0("The prespecified abiotic–spatial model improved WAIC by ",
         fmt(m0_waic - m1_waic), " relative to the spatial-only model."),
  paste0("Adding the exploratory Bombus-associated index changed WAIC by ",
         fmt(m2_waic - m1_waic),
         " relative to the matched abiotic–spatial model. Anomalies were calculated from the minimum-WAIC selected model."),
  paste0("Abiotic predictors with 95% credible intervals excluding zero were: ",
         if (length(supported_terms)) paste(supported_terms, collapse = ", ") else "none", "."),
  paste0(sum(dat_used$upper_anomaly_95, na.rm = TRUE),
         " observations exceeded the upper 95% leave-one-out predictive threshold.")
)
writeLines(results_lines, file.path(text_dir, "results_key_sentences.txt"))

caption_lines <- c(
  "Figure 1. Georeferenced phenotyping and continuous floral-colour variation. Photograph-derived median sRGB values were converted to CIELAB and summarized by a prespecified latent display-colour model.",
  "Figure 2. Fixed effects from the minimum-WAIC ecological–spatial model and support for the prespecified M0–M2 model set. Filled points have 95% credible intervals excluding zero; models with ΔWAIC < 2 have comparable support.",
  "Figure 3. Exploratory Bombus-associated context. The left panel assesses dependence on the suitability-composite definition; the right panel shows overlap between the primary Bombus index and measured environmental axes.",
  "Figure 4. Selected-model anomalies. Positive PIT-normal anomalies indicate flowers darker than predicted by the minimum-WAIC ecological–spatial model in approximate leave-one-out prediction. Upper-tail enrichment and quantile relationships are exploratory and do not establish horticultural origin."
)
writeLines(caption_lines, file.path(text_dir, "figure_captions.txt"))

review_scope <- c(
  "Confirmatory core: photograph-derived continuous floral colour modelled against prespecified abiotic domains with an SPDE spatial field.",
  "Exploratory biotic extension: species-specific, spatially cross-validated Bombus suitability SDMs; suitability is not occurrence probability, abundance, visitation, or selection.",
  "Exploratory anomaly stage: upper-tail departure from the selected ecological–spatial model; association with human exposure does not establish horticultural origin.",
  "The SPDE field represents residual spatial dependence and cannot be uniquely attributed to history, genetics, or unmeasured environment.",
  "PCA reduces within-domain dimension; it does not make ecological domains or the Bombus index independent."
)
writeLines(review_scope, file.path(text_dir, "scope_and_inference_limits.txt"))

analysis_config <- list(
  pipeline_version = "FINAL_PUBLICATION_CLEAN_20KM",
  run_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
  random_seed = 42,
  input_csv = normalizePath(path.expand(in_csv), winslash = "/", mustWork = TRUE),
  input_csv_md5 = unname(tools::md5sum(path.expand(in_csv))),
  rmd_path = tryCatch(normalizePath(knitr::current_input(), winslash = "/", mustWork = TRUE), error = function(e) NA_character_),
  rmd_md5 = tryCatch(unname(tools::md5sum(knitr::current_input())), error = function(e) NA_character_),
  output_root = normalizePath(out_root, winslash = "/", mustWork = TRUE),
  selected_model = best_model_name,
  primary_models = primary_terms,
  environment_terms = environment_terms,
  bombus_species = bombus_species,
  bombus_background_n = BOMBUS_BACKGROUND_N,
  bombus_coordinate_uncertainty_max_m = BOMBUS_MAX_COORD_UNCERTAINTY_M,
  mesh = list(max_edge_m = c(20000, 100000), cutoff_m = 5000),
  qgam_quantiles = taus
)
jsonlite::write_json(analysis_config,
                     file.path(repro_dir, "analysis_config.json"),
                     pretty = TRUE, auto_unbox = TRUE, null = "null")

saveRDS(
  list(
    fit_pigment_cfa = fit_pig,
    pca = list(
      temperature = temp_pc_obj,
      moisture = precip_pc_obj,
      soil_physical = soil_phys_pc_obj,
      soil_nutrient = soil_nutrient_pc_obj,
      topography = topo_pc_obj
    ),
    primary_fits = primary_fits,
    selected_model_name = best_model_name,
    qgam_models = qgam_models
  ),
  file.path(repro_dir, "fitted_analysis_objects.rds"),
  compress = "xz"
)

capture.output(sessionInfo(), file = file.path(repro_dir, "sessionInfo.txt"))
installed_used <- unique(c(
  "dplyr", "readr", "tidyr", "sf", "terra", "lavaan", "tibble", "INLA",
  "qgam", "mgcv", "ggplot2", "forcats", "purrr", "broom", "jsonlite",
  "digest", required_sdm_pkgs
))
package_versions <- tibble::tibble(
  package = installed_used,
  version = vapply(installed_used, function(pkg) {
    if (requireNamespace(pkg, quietly = TRUE)) as.character(utils::packageVersion(pkg)) else NA_character_
  }, character(1))
)
write_csv(package_versions, file.path(repro_dir, "package_versions.csv"))

all_outputs <- list.files(pub_dir, recursive = TRUE, full.names = TRUE)
all_outputs <- all_outputs[file.info(all_outputs)$isdir %in% FALSE]
output_manifest <- tibble::tibble(
  file = sub(paste0("^", normalizePath(pub_dir, winslash = "/"), "/?"), "",
             normalizePath(all_outputs, winslash = "/", mustWork = TRUE)),
  bytes = file.info(all_outputs)$size,
  md5 = unname(tools::md5sum(all_outputs))
)
write_csv(output_manifest, file.path(repro_dir, "publication_output_manifest.csv"))
```

## Complete

```{r completion}
cat("Submission materials:", pub_dir, "\n")
cat("Reproducibility archive:", repro_dir, "\n")
```
