---
title: "Campanula punctata floral-colour geography: phenotype analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = TRUE, message = FALSE, fig.align = "center", dpi = 150
)

hurdle_dir <- Sys.getenv(
  "HOTARUBUKURO_HURDLE_RESULTS",
  "results/ecological_v11_pigmentation_hurdle"
)
base_dir <- Sys.getenv(
  "HOTARUBUKURO_BASE_RESULTS",
  "results/ecological_v9_final_public_HRNA_50km"
)

read_required <- function(dir, name) {
  path <- file.path(dir, name)
  if (!file.exists(path)) stop("Missing required phenotype result: ", path)
  utils::read.csv(path, check.names = FALSE, stringsAsFactors = FALSE)
}

measurement <- read_required(hurdle_dir, "pigmentation_measurement_summary.csv")
components <- read_required(hurdle_dir, "pigmentation_mixture_components.csv")
measurement_obs <- read_required(hurdle_dir, "pigmentation_measurement_observations.csv")
class_sensitivity <- read_required(hurdle_dir, "pigmentation_classification_sensitivity.csv")
presence_coef <- read_required(hurdle_dir, "pigmentation_presence_bombus_coefficients.csv")
presence_held <- read_required(hurdle_dir, "pigmentation_presence_bombus_heldout.csv")
intensity_coef <- read_required(hurdle_dir, "pigmented_intensity_bombus_coefficients.csv")
intensity_held <- read_required(hurdle_dir, "pigmented_intensity_bombus_heldout.csv")
rule_coef <- read_required(hurdle_dir, "pigmentation_rule_sensitivity_coefficients.csv")
rule_held <- read_required(hurdle_dir, "pigmentation_rule_sensitivity_heldout.csv")
inla_cmp <- read_required(hurdle_dir, "pigmentation_hurdle_inla_model_comparison.csv")
inla_fixed <- read_required(hurdle_dir, "pigmentation_hurdle_inla_fixed_effects.csv")
hr_coef <- read_required(hurdle_dir, "pigmentation_hurdle_HR_coefficients.csv")
hr_held <- read_required(hurdle_dir, "pigmentation_hurdle_HR_heldout.csv")
tail_coef <- read_required(hurdle_dir, "pigmentation_residual_tail_HR_coefficients.csv")
tail_held <- read_required(hurdle_dir, "pigmentation_residual_tail_HR_heldout.csv")
tail_early <- read_required(hurdle_dir, "pigmentation_residual_tail_early_enrichment.csv")
tail_candidates <- read_required(hurdle_dir, "pigmentation_residual_tail_candidates.csv")
vif <- read_required(hurdle_dir, "pigmentation_hurdle_collinearity_vif.csv")
condition <- read_required(hurdle_dir, "pigmentation_hurdle_collinearity_condition.csv")
enmeval <- read_required(base_dir, "bombus_enmeval_selection.csv")

num <- function(x, digits = 3) formatC(as.numeric(x), digits = digits, format = "f")
one <- function(x, keep, label) {
  out <- x[keep, , drop = FALSE]
  if (nrow(out) != 1L) stop("Expected one row for ", label, "; found ", nrow(out))
  out
}
coef_ci <- function(row) paste0(
  num(row$estimate), " (95% interval ", num(row$lower_95), " to ",
  num(row$upper_95), ")"
)
inla_ci <- function(row) paste0(
  num(row$mean), " (95% credible interval ", num(row[["0.025quant"]]),
  " to ", num(row[["0.975quant"]]), ")"
)

w_presence <- one(
  presence_coef,
  presence_coef$predictor == "Bombus_W" &
    presence_coef$rung == "environment_space_adjusted",
  "W presence env-space coefficient"
)
a_presence <- one(
  presence_coef,
  presence_coef$predictor == "Bombus_A" &
    presence_coef$rung == "environment_space_adjusted",
  "A presence env-space coefficient"
)
w_intensity <- one(
  intensity_coef,
  intensity_coef$predictor == "Bombus_W" &
    intensity_coef$rung == "environment_space_adjusted",
  "W intensity env-space coefficient"
)
a_intensity <- one(
  intensity_coef,
  intensity_coef$predictor == "Bombus_A" &
    intensity_coef$rung == "environment_space_adjusted",
  "A intensity env-space coefficient"
)

w_presence_cv <- one(
  presence_held,
  presence_held$predictor == "Bombus_W" &
    presence_held$rung == "environment_space_adjusted",
  "W presence env-space heldout"
)
a_presence_cv <- one(
  presence_held,
  presence_held$predictor == "Bombus_A" &
    presence_held$rung == "environment_space_adjusted",
  "A presence env-space heldout"
)
w_intensity_cv <- one(
  intensity_held,
  intensity_held$predictor == "Bombus_W" &
    intensity_held$rung == "environment_space_adjusted",
  "W intensity env-space heldout"
)
a_intensity_cv <- one(
  intensity_held,
  intensity_held$predictor == "Bombus_A" &
    intensity_held$rung == "environment_space_adjusted",
  "A intensity env-space heldout"
)
```

# Scope and result

This stage report follows the publication design in `docs/analysis-plan.md`.
It replaces the superseded single continuous flower-colour response with a
two-part pigmentation response. Old monolithic-workflow results are not carried
forward.

The defensible result is:

> Author-confirmed photograph regions revealed a nationwide optical transition
> between white-like and pigmented flowers and a separate intensity gradient
> among pigmented flowers. Bombus suitability overlaps both gradients at broad
> spatial scales, but pigmentation presence gains essentially no predictive
> information from W or A after environmental and spatial structure is included.
> A small W-associated intensity gradient remains among pigmented flowers, but
> it does not establish pollinator-mediated selection. Human--forest interface
> associations are weak and do not become stronger in the extreme residual
> tail; horticultural provenance therefore remains an unconfirmed candidate
> explanation rather than a supported population-level result.

# Measurement: two biological regimes

For each photograph, the focal floral subject was extracted/cropped and the
author visually verified the retained petal colour-extraction region before
scripted colour calculation. Mask shape was not used to decide biological
eligibility. Extreme dark or pink values were retained because they can be real
flowers and are central to the horticultural hypothesis.

White flowers do not provide a continuous proxy for anthocyanin amount. Their
small a* differences are therefore treated as colour-measurement variation.
The primary response-blind Gaussian mixture fitted only to a* estimated an
operational boundary of **a* = `r num(measurement$decision_boundary_a, 2)`**:
`r measurement$n_white` white-like and `r measurement$n_pigmented` pigmented
observations. `r measurement$n_ambiguous` observations fall in the posterior
uncertainty zone at an 0.8 confidence rule.

The two outcomes are:

1. pigmentation presence/absence across all `r measurement$n` flowers;
2. standardized visible colour intensity among the `r measurement$n_pigmented`
   flowers classified as pigmented.

These are optical phenotypes from uncalibrated display-referred sRGB converted
to CIELAB. They are not chemical anthocyanin concentrations, reflectance spectra,
or bee-perceptual contrasts.

```{r mixture-figure, fig.width=8, fig.height=4.5, fig.cap="Observed CIELAB a* distribution and the response-blind operational pigmentation boundary."}
hist(
  measurement_obs$colour_a, breaks = 60, col = "grey85", border = "white",
  xlab = "CIELAB a*", main = "White-like and pigmented measurement regimes"
)
abline(v = measurement$decision_boundary_a, col = "#B2182B", lwd = 3)
legend(
  "topright", legend = paste0("mixture boundary = ", num(measurement$decision_boundary_a, 2)),
  lwd = 3, col = "#B2182B", bty = "n"
)
```

```{r classification-table}
knitr::kable(
  class_sensitivity, digits = 3,
  caption = "Pigmentation classification by region under prespecified sensitivity rules."
)
```

The rule a*>0 is not the primary boundary: it classifies
`r measurement$n_nonnegative_below_boundary` non-negative but white-like
observations as pigmented. High-confidence exclusion and the joint L*a*C*
mixture are the biologically more informative robustness checks.

# Analysis architecture and anti-circularity

The same causal idea is evaluated as a sequence of distinct estimands rather
than as one saturated model.

1. **Natural baseline:** eight response-blind environmental axes plus region and
   a continuous spatial field.
2. **Biotic addition:** W nationwide; A only within its common ENMeval support.
3. **Horticultural increment:** H (human exposure), R (human--forest interface),
   and H×R are added to the original outcome model. This is the inferential test.
4. **Residual-tail exploration:** a natural model that contains no H, R, road
   access, early-flowering score, or residual rank produces spatially
   cross-fitted predictions. Its residuals rank unexpected observations only.

Residuals are therefore not treated as a new measured trait, and candidate
selection is not recycled as confirmation of the H/R model.

# Bombus: broad overlap versus incremental information

## ENMeval suitability inputs

All five suitability models were rebuilt and selected with ENMeval; existing
Bombus TIFF files were not used as trusted inputs. AICc, not AUC, selected model
complexity. The broad-distribution B. diversus model has modest validation AUC,
which is expected when presence locations occupy much of the accessible
background; it limits fine-scale discrimination and is reported rather than
hidden.

```{r enmeval-table}
enmeval_show <- enmeval[c(
  "species", "fc", "rm", "auc.val.avg", "auc.val.sd", "cbi.val.avg",
  "or.10p.avg", "AICc", "ncoef"
)]
knitr::kable(enmeval_show, digits = 3, caption = "Selected ENMeval models.")
```

## Evidence ladder

At the landscape rung, W is negatively associated and A positively associated
with pigmentation presence. After environment and space, the W coefficient is
`r coef_ci(w_presence)` and the A coefficient is `r coef_ci(a_presence)`.
Their spatial-fold predictive changes are respectively
Δlog-loss=`r num(w_presence_cv$delta_log_loss, 4)` and
`r num(a_presence_cv$delta_log_loss, 4)`: effectively no gain.

Among pigmented flowers, W retains a negative coefficient,
`r coef_ci(w_intensity)`, with Δheld-out R²=
`r num(w_intensity_cv$delta_R2, 4)`. A is `r coef_ci(a_intensity)` with
Δheld-out R²=`r num(a_intensity_cv$delta_R2, 4)`.

```{r bombus-ladder-table}
coef_show <- rbind(
  subset(presence_coef, predictor %in% c("Bombus_W", "Bombus_A") &
           rung %in% c("landscape", "environment_space_adjusted")),
  subset(intensity_coef, predictor %in% c("Bombus_W", "Bombus_A") &
           rung %in% c("landscape", "environment_space_adjusted"))
)
knitr::kable(
  coef_show[c("outcome", "predictor", "rung", "n", "estimate", "lower_95", "upper_95", "p_value")],
  digits = 4, caption = "Bombus evidence ladder for the two response components."
)
```

The three montane species show similar broad positive overlap but are highly
correlated spatial proxies, not independent biological replications. One
B. consobrinus presence cross-validation fold emitted an mgcv step-failure
warning; the W/A results and all conditional-intensity folds had no such warning.

## Threshold robustness

The primary conclusion is stable under high-confidence classification and the
joint L*a*C* mixture. The a*>0 rule alone makes the adjusted A coefficient look
positive, demonstrating why including white-like a* noise as low pigment can
change inference.

```{r threshold-robustness-table}
rule_show <- subset(
  rule_coef,
  rung == "environment_space_adjusted" & predictor %in% c("Bombus_W", "Bombus_A")
)
knitr::kable(
  rule_show[c(
    "classification_rule", "outcome_role", "predictor", "n", "estimate",
    "lower_95", "upper_95", "p_value"
  )],
  digits = 4, caption = "Environment-and-space-adjusted threshold sensitivity."
)
```

# SPDE-INLA check

The binary presence outcome uses a binomial-logit SPDE model; conditional
intensity uses a Gaussian SPDE model. Both use the same mesh construction and
PC-Matérn prior specification. Adding W to the nationwide presence model worsens
WAIC by 1.37. Adding A within common support worsens WAIC by 2.71. For intensity,
W improves nationwide WAIC by only 0.54 and A changes it by 0.04.

```{r inla-comparison-table}
knitr::kable(
  inla_cmp[c(
    "model", "family", "n", "mesh_vertices", "WAIC", "mean_neglogCPO",
    "n_CPO_nonfinite", "comparison_set", "delta_WAIC_within_set"
  )],
  digits = 3, caption = "Two-part SPDE-INLA model comparison."
)
```

# Shared structure and multicollinearity

Environment-only axes are not the main identifiability problem. Bombus W is
strongly aligned with temperature, especially in the pigmented and eastern
subsets. Within A support, W and A are also strongly negatively correlated.
This is expected shared biogeography, but it widens intervals and prevents a
clean local selection interpretation.

```{r collinearity-table}
max_vif <- do.call(rbind, lapply(split(vif, vif$design), function(x) {
  x[which.max(x$VIF), c("design", "n", "term", "VIF")]
}))
knitr::kable(max_vif, digits = 3, caption = "Maximum VIF in each actual analysis subset.")
knitr::kable(
  condition[c("design", "n", "p", "condition_number")], digits = 3,
  caption = "Condition numbers in the same subsets."
)
```

Variables are not automatically deleted by a VIF threshold because W, A, H and
R represent prespecified processes. Instead, the claim is bounded by intervals,
WAIC/CPO, spatial-fold prediction and proxy sensitivities. In the most difficult
intensity A-support design, the maximum VIF is about 41 and the condition number
about 18.3; W-versus-A attribution there is consequently weak.

# Horticultural hypothesis

## Direct incremental test

In western Japan, the direct presence model gives a positive H×R coefficient,
but the effect is small in held-out prediction. Conditional intensity has a
positive point estimate with an interval crossing zero. Date, access and mesh-
boundary sensitivities do not turn this into strong convergent evidence.

```{r horticulture-direct-table}
direct_west <- subset(
  hr_coef, region == "West" & term == "z_H_R"
)
direct_cv_west <- subset(
  hr_held, region == "West" & increment == "H_by_R_only"
)
knitr::kable(
  direct_west[c(
    "outcome", "variant", "n", "events", "estimate", "lower_95", "upper_95", "p_value"
  )],
  digits = 4, caption = "Western H-by-R coefficients on the original two-part outcomes."
)
knitr::kable(
  direct_cv_west[c(
    "outcome", "variant", "n", "delta_log_loss", "delta_auc", "delta_R2",
    "base_R2", "full_R2"
  )],
  digits = 4, caption = "Spatial-fold incremental performance for H-by-R."
)
```

## Cross-fitted residual-tail test

The residual analysis asks a narrower question: after a natural model is fitted
without human variables, do H and R identify increasingly extreme departures?
The answer is no. Western unexpected presence has a signal at the 80th
percentile, but it disappears at 90% and 95%. For intensity, H×R changes sign,
all intervals cross zero, and every western threshold worsens spatial-fold
log-loss.

```{r residual-tail-table}
tail_west <- subset(
  tail_coef, region == "West" & variant == "primary" & term == "z_H_R"
)
tail_cv_west <- subset(tail_held, region == "West" & variant == "primary")
knitr::kable(
  tail_west[c(
    "outcome", "threshold", "n", "events", "estimate", "lower_95", "upper_95", "p_value"
  )],
  digits = 4, caption = "Western H-by-R estimates across cross-fitted residual tails."
)
knitr::kable(
  tail_cv_west[c(
    "outcome", "threshold", "n", "delta_log_loss", "delta_auc", "n_warning_folds"
  )],
  digits = 4, caption = "Held-out performance across the same residual tails."
)
```

Early flowering also fails as a western convergent signature. Among the eight
western flowers in the top 5% conditional-intensity residual tail, none belongs
to the early-score top 10%. This does not prove natural origin for individual
flowers; it falsifies the current population-level early-cultivar prediction.

```{r early-table}
knitr::kable(
  tail_early, digits = 3,
  caption = "Early-flowering enrichment in cross-fitted residual tails."
)
```

## Candidate ranking is not provenance

The candidate table remains useful for follow-up because it identifies flowers
that the natural model predicts poorly. Ranking uses cross-fitted residuals; it
does not label horticultural origin. Exact locality is omitted here.

```{r candidate-table}
candidate_show <- subset(
  tail_candidates,
  candidate_outcome == "pigmented_intensity" & region == "West"
)
candidate_show <- head(candidate_show[order(-candidate_show$candidate_percentile), ], 12)
knitr::kable(
  candidate_show[c(
    "observation_id", "candidate_percentile", "colour_a",
    "natural_intensity_prediction", "intensity_natural_residual", "z_H", "z_R",
    "early_score", "z_A", "DOY", "year", "spatial_fold"
  )],
  digits = 3, caption = "Top western conditional-intensity residual candidates for external validation."
)
```

# Reviewer-facing interpretation

| Question | Supported statement | Not supported |
|---|---|---|
| Colour geography | Optical pigmentation presence and conditional intensity have broad environmental and spatial gradients. | a* in white flowers is anthocyanin quantity. |
| Bombus | W/A guild turnover overlaps flower pigmentation geography; a small W-associated conditional-intensity gradient remains. | A local causal pollinator-selection effect independent of shared climate and geography. |
| Space | SPDE quantifies structured residual geography and shows that spatial structure is biologically informative. | The spatial field is a specific historical or genetic mechanism. |
| Horticulture | Human-interface variables can rank candidates and show a weak direct western presence association. | Residual extremes become increasingly horticultural, or horticultural provenance is established. |

The main novelty is therefore methodological and system-specific: large-scale
SNS phenotyping can separate pigmentation expression from intensity, explicitly
partition environmental and spatial structure, and then evaluate a biotic proxy
and human-interface hypothesis without treating either collinearity or residuals
as a shortcut to causality. Campanula punctata is especially informative because
its known Bombus pollination, mainland colour polymorphism, white-flowered island
relative in a Bombus-poor setting, and horticultural use generate competing
predictions that can be made explicit.

# Limitations and next validation

- The optical threshold lacks chemical anthocyanin or reflectance validation.
- Occurrence/suitability proxies are not visitation, abundance or selection
  gradients, and the broad species are difficult to discriminate spatially.
- H, R and road access are exposure proxies, not records of planting or escape.
- Cross-fitted residuals reduce fitting circularity but remain model-dependent.
- Exact provenance requires cultivar genotypes, planting records, field
  observations, or targeted sampling of the ranked candidates.

The current paper can lead with the two-part colour geography and SPDE result,
present Bombus as a carefully bounded biotic extension, and present horticulture
as a falsifiable exploratory workflow whose strongest predicted convergence was
not observed.
