---
title: "Flower color pipeline"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## FINAL PIPELINE

### 0. Packages and reproducibility settings

```{r packages}
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(tidyr)
  library(sf)
  library(terra)
  library(lavaan)
  library(tibble)
  library(INLA)
  library(qgam)
  library(mgcv)
  library(ggplot2)
  library(forcats)
  library(purrr)
})

set.seed(42)
INLA::inla.setOption(num.threads = "1:1")

if (requireNamespace("RhpcBLASctl", quietly = TRUE)) {
  RhpcBLASctl::blas_set_num_threads(1)
  RhpcBLASctl::omp_set_num_threads(1)
}

RNGkind(kind = "Mersenne-Twister", normal.kind = "Inversion")

z <- function(x) as.numeric(scale(x))

select <- dplyr::select
filter <- dplyr::filter
mutate <- dplyr::mutate
across <- dplyr::across

out_root <- path.expand("~/zui_results")
dir.create(out_root, recursive = TRUE, showWarnings = FALSE)

cmp_dir      <- file.path(out_root, "INLA_auto_competing_pair")
top_dir      <- file.path(out_root, "INLA_top_models_for_supp")
best_dir     <- file.path(out_root, "INLA_best_model_final")
best_plotdir <- file.path(best_dir, "plots")
gam_dir      <- file.path(out_root, "qGAM_best_model_final")

dir.create(cmp_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(top_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(best_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(best_plotdir, recursive = TRUE, showWarnings = FALSE)
dir.create(gam_dir, recursive = TRUE, showWarnings = FALSE)
```

###1. Load CSV and compute color variables

```{r load-color}
in_csv <- "~/Desktop/1006.csv"
df <- read.csv(in_csv, stringsAsFactors = FALSE)

stopifnot(all(c("R", "G", "B") %in% names(df)))
stopifnot(all(c("longitude", "latitude") %in% names(df)))

if (!("sample_id" %in% names(df))) {
  df$sample_id <- seq_len(nrow(df))
}

## HSV
hsv_mat <- t(grDevices::rgb2hsv(df$R, df$G, df$B))
colnames(hsv_mat) <- c("H", "S", "V")
df <- cbind(df, as.data.frame(hsv_mat))

## Lab
rgb01 <- pmin(pmax(as.matrix(df[, c("R", "G", "B")]) / 255, 0), 1)
lab_mat <- grDevices::convertColor(rgb01, from = "sRGB", to = "Lab")
df$L  <- lab_mat[, 1]
df$a  <- lab_mat[, 2]
df$b  <- lab_mat[, 3]
df$C  <- sqrt(df$a^2 + df$b^2)
df$Lm <- -df$L

## hex
rng_rgb <- range(c(df$R, df$G, df$B), na.rm = TRUE)
is_01 <- (rng_rgb[1] >= 0 && rng_rgb[2] <= 1)

R <- df$R
G <- df$G
B <- df$B
if (is_01) {
  R <- pmin(pmax(round(R * 255), 0), 255)
  G <- pmin(pmax(round(G * 255), 0), 255)
  B <- pmin(pmax(round(B * 255), 0), 255)
}
df$col_hex <- rgb(R, G, B, maxColorValue = 255)

cat("✓ Loaded & color features computed. n =", nrow(df), "\n")
```

###  Extract raster values

```{r raster-extraction}
chelsa_dir <- "~/Desktop/stress2"
soil_dir   <- "~/Desktop/soil2"
bee_dir    <- "~/Desktop/bee5"
human_dir  <- "~/Desktop/human"

coords <- df %>%
  select(longitude, latitude) %>%
  mutate(across(everything(), as.numeric)) %>%
  filter(complete.cases(.))

extract_folder_terra <- function(path, coords) {
  files <- list.files(path, pattern = "\\.tif$", full.names = TRUE)
  if (!length(files)) return(NULL)

  pts <- terra::vect(coords, geom = c("longitude", "latitude"), crs = "EPSG:4326")

  vals_list <- lapply(files, function(f) {
    r <- terra::rast(f)
    pts_use <- if (!terra::same.crs(pts, r)) terra::project(pts, terra::crs(r)) else pts
    terra::extract(r, pts_use, ID = FALSE)[, 1]
  })

  out <- as.data.frame(vals_list)
  names(out) <- tools::file_path_sans_ext(basename(files))
  out
}

chelsa_vals <- extract_folder_terra(chelsa_dir, coords)
soil_vals   <- extract_folder_terra(soil_dir, coords)
bee_vals    <- extract_folder_terra(bee_dir, coords)

n_row <- nrow(coords)
if (!is.null(chelsa_vals)) stopifnot(nrow(chelsa_vals) == n_row)
if (!is.null(soil_vals))   stopifnot(nrow(soil_vals)   == n_row)
if (!is.null(bee_vals))    stopifnot(nrow(bee_vals)    == n_row)

df <- cbind(df[seq_len(n_row), ], chelsa_vals, soil_vals, bee_vals)

df2 <- df %>% distinct(longitude, latitude, .keep_all = TRUE)
cat("✓ Raster extracted. df2 n =", nrow(df2), "\n")
```

### 3. Derive DEM-based topographic variables

```{r dem-topography}
dem_file <- file.path(soil_dir, "elevation.tif")

if (file.exists(dem_file)) {
  dem_r <- terra::rast(dem_file)

  slope_r  <- terra::terrain(dem_r, v = "slope",  unit = "radians")
  aspect_r <- terra::terrain(dem_r, v = "aspect", unit = "radians")
  tri_r    <- terra::terrain(dem_r, v = "TRI")
  tpi_r    <- terra::terrain(dem_r, v = "TPI")

  names(slope_r)  <- "slope"
  names(aspect_r) <- "aspect"
  names(tri_r)    <- "TRI"
  names(tpi_r)    <- "TPI"

  topo_stack <- c(slope_r, aspect_r, tri_r, tpi_r)

  pts <- terra::vect(
    df2 %>% dplyr::select(longitude, latitude),
    geom = c("longitude", "latitude"),
    crs  = "EPSG:4326"
  )

  pts_use <- if (!terra::same.crs(pts, topo_stack)) {
    terra::project(pts, terra::crs(topo_stack))
  } else {
    pts
  }

  topo_vals <- terra::extract(topo_stack, pts_use, ID = FALSE)
  df2 <- cbind(df2, topo_vals)

  df2$northness <- cos(df2$aspect)
  df2$eastness  <- sin(df2$aspect)

  cat("✓ DEM-derived topographic variables added from:", dem_file, "\n")
} else {
  cat("! DEM file not found:", dem_file, "\n")
}
```

### 4. Run PCA for environmental predictor sets

```{r pca}
set.seed(42)

## Temperature PCA
temp_vars <- grep(
  "^(CHELSA_bio5_1981-2010_V\\.2\\.1|CHELSA_bio10_1981-2010_V\\.2\\.1|CHELSA_gdd5_1981-2010_V\\.2\\.1)$",
  names(df2), value = TRUE
)

Temperature_PC1 <- rep(NA_real_, nrow(df2))
temp_pc_obj <- NULL

if (length(temp_vars) >= 1) {
  Xt <- df2[, temp_vars, drop = FALSE]
  cc <- complete.cases(Xt)
  if (sum(cc) >= 3 && ncol(Xt) >= 2) {
    temp_pc_obj <- prcomp(Xt[cc, ], scale. = TRUE)
    Temperature_PC1[cc] <- temp_pc_obj$x[, 1]
  } else if (ncol(Xt) == 1) {
    Temperature_PC1 <- as.numeric(scale(Xt[[1]]))
  }
}
df2$Temperature_PC1 <- Temperature_PC1

## Precipitation / dryness PCA
precip_vars <- grep(
  "^(CHELSA_ai_1981-2010_V\\.2\\.1|CHELSA_vpd_mean_1981-2010_V\\.2\\.1|CHELSA_bio1[245]_1981-2010_V\\.2\\.1|CHELSA_swb_1981-2010_V\\.2\\.1)$",
  names(df2), value = TRUE
)

precip_PC1 <- rep(NA_real_, nrow(df2))
precip_pc_obj <- NULL

if (length(precip_vars) >= 1) {
  Xp <- df2[, precip_vars, drop = FALSE]
  cc <- complete.cases(Xp)
  if (sum(cc) >= 3 && ncol(Xp) >= 2) {
    precip_pc_obj <- prcomp(Xp[cc, ], scale. = TRUE)
    precip_PC1[cc] <- precip_pc_obj$x[, 1]
  } else if (ncol(Xp) == 1) {
    precip_PC1 <- as.numeric(scale(Xp[[1]]))
  }
}
df2$precip_PC1 <- precip_PC1

## Soil physical PCA
soil_phys_vars <- grep(
  "^(bdod_0[.-]5cm_mean|cfvo_0[.-]5cm_mean|sand_0[.-]5cm_mean|silt_0[.-]5cm_mean)$",
  names(df2), value = TRUE
)

soil_phys_PC1 <- rep(NA_real_, nrow(df2))
soil_phys_pc_obj <- NULL

if (length(soil_phys_vars) >= 1) {
  Xp <- df2[, soil_phys_vars, drop = FALSE]
  keep <- vapply(Xp, function(v) is.numeric(v) && sd(v, na.rm = TRUE) > 0, logical(1))
  Xp <- Xp[, keep, drop = FALSE]
  if (ncol(Xp) >= 2) {
    cc <- complete.cases(Xp)
    if (sum(cc) >= 3) {
      soil_phys_pc_obj <- prcomp(Xp[cc, ], scale. = TRUE)
      soil_phys_PC1[cc] <- soil_phys_pc_obj$x[, 1]
    }
  } else if (ncol(Xp) == 1) {
    soil_phys_PC1 <- as.numeric(scale(Xp[[1]]))
  }
}
df2$soil_phys_PC1 <- soil_phys_PC1

## Soil nutrient PCA
soil_nutrient_vars <- grep(
  "^(nitrogen_0[.-]5cm_mean|ocd_0[.-]5cm_mean|soc_0[.-]5cm_mean)$",
  names(df2), value = TRUE
)

soil_nutrient_PC1 <- rep(NA_real_, nrow(df2))
soil_nutrient_pc_obj <- NULL

if (length(soil_nutrient_vars) >= 1) {
  Xn <- df2[, soil_nutrient_vars, drop = FALSE]
  keep <- vapply(Xn, function(v) is.numeric(v) && sd(v, na.rm = TRUE) > 0, logical(1))
  Xn <- Xn[, keep, drop = FALSE]
  if (ncol(Xn) >= 2) {
    cc <- complete.cases(Xn)
    if (sum(cc) >= 3) {
      soil_nutrient_pc_obj <- prcomp(Xn[cc, ], scale. = TRUE)
      soil_nutrient_PC1[cc] <- soil_nutrient_pc_obj$x[, 1]
    }
  } else if (ncol(Xn) == 1) {
    soil_nutrient_PC1 <- as.numeric(scale(Xn[[1]]))
  }
}
df2$soil_nutrient_PC1 <- soil_nutrient_PC1

## Soil pH
soil_pH_var <- grep("^phh2o_0[.-]5cm_mean$", names(df2), value = TRUE)
if (length(soil_pH_var) >= 1) {
  df2$soil_pH <- df2[[soil_pH_var[1]]]
} else {
  df2$soil_pH <- NA_real_
}

## Topography PCA
topo_vars <- intersect(c("roughness", "slope", "TRI"), names(df2))

topo_PC1 <- rep(NA_real_, nrow(df2))
topo_pc_obj <- NULL

if (length(topo_vars) >= 2) {
  Xt <- df2[, topo_vars, drop = FALSE]
  keep <- vapply(Xt, function(v) is.numeric(v) && sd(v, na.rm = TRUE) > 0, logical(1))
  Xt <- Xt[, keep, drop = FALSE]
  if (ncol(Xt) >= 2) {
    cc <- complete.cases(Xt)
    if (sum(cc) >= 3) {
      topo_pc_obj <- prcomp(Xt[cc, ], scale. = TRUE)
      topo_PC1[cc] <- topo_pc_obj$x[, 1]
    }
  } else if (ncol(Xt) == 1) {
    topo_PC1 <- as.numeric(scale(Xt[[1]]))
  }
}
df2$topo_PC1 <- topo_PC1

cat("✓ PCA done.\n")
```

### 6. CFA Pigment latent factor

```{r cfa-pigment}
stopifnot(all(c("a","Lm","C") %in% names(df2)))

model_pig <- ' Pigment =~ 1*a + Lm + C '
fit_pig <- lavaan::cfa(
  model_pig,
  data      = df2[, c("a","Lm","C")],
  estimator = "MLR",
  std.lv    = FALSE,
  missing   = "fiml"
)

stopifnot(lavaan::inspect(fit_pig, "converged"))
df2$Pigment <- as.numeric(lavaan::lavPredict(fit_pig, type = "lv")[, "Pigment"])

cat("✓ Pigment factor computed.\n")
```

### 5. Export PCA loadings

```{r export-loadings}
if (!is.null(temp_pc_obj)) {
  write_csv(
    as.data.frame(temp_pc_obj$rotation) %>% tibble::rownames_to_column("variable"),
    file.path(cmp_dir, "temperature_pca_loadings.csv")
  )
}
if (!is.null(precip_pc_obj)) {
  write_csv(
    as.data.frame(precip_pc_obj$rotation) %>% tibble::rownames_to_column("variable"),
    file.path(cmp_dir, "precip_pca_loadings.csv")
  )
}
if (!is.null(soil_phys_pc_obj)) {
  write_csv(
    as.data.frame(soil_phys_pc_obj$rotation) %>% tibble::rownames_to_column("variable"),
    file.path(cmp_dir, "soil_physical_pca_loadings.csv")
  )
}
if (!is.null(soil_nutrient_pc_obj)) {
  write_csv(
    as.data.frame(soil_nutrient_pc_obj$rotation) %>% tibble::rownames_to_column("variable"),
    file.path(cmp_dir, "soil_nutrient_pca_loadings.csv")
  )
}
if (!is.null(topo_pc_obj)) {
  write_csv(
    as.data.frame(topo_pc_obj$rotation) %>% tibble::rownames_to_column("variable"),
    file.path(cmp_dir, "topography_pca_loadings.csv")
  )
}
```

### 6. Build RSDS, pigment factor, and bee richness

```{r rsds-pigment-bee}
rsds_candidates <- grep("rsds", names(df2), value = TRUE, ignore.case = TRUE)
stopifnot(length(rsds_candidates) > 0)
df2$RSDS <- df2[[rsds_candidates[1]]]

stopifnot(all(c("a", "Lm", "C") %in% names(df2)))
model_pig <- ' Pigment =~ 1*a + Lm + C '
fit_pig <- lavaan::cfa(
  model_pig,
  data      = df2[, c("a", "Lm", "C")],
  estimator = "MLR",
  std.lv    = FALSE,
  missing   = "fiml"
)
stopifnot(lavaan::inspect(fit_pig, "converged"))
df2$Pigment <- as.numeric(lavaan::lavPredict(fit_pig, type = "lv")[, "Pigment"])

bee_cols <- intersect(c("ardens", "beaticola", "consobrinus", "diversus", "honshuensis"), names(df2))
stopifnot(length(bee_cols) > 0)
df2$Bee_Richness <- rowSums(df2[, bee_cols], na.rm = TRUE)

cat("✓ RSDS / Pigment / Bee richness ready.\n")
```

### 7. Build modeling dataset

```{r build-model-data}
need_vars <- c(
  "Pigment",
  "topo_PC1",
  "Temperature_PC1",
  "precip_PC1",
  "soil_nutrient_PC1",
  "soil_phys_PC1",
  "soil_pH",
  "RSDS",
  "Bee_Richness",
  "longitude", "latitude"
)

dat0 <- df2 %>%
  dplyr::select(all_of(need_vars), any_of("date"), any_of("elevation")) %>%
  dplyr::filter(complete.cases(across(all_of(need_vars)))) %>%
  dplyr::mutate(
    y         = Pigment,
    z_RSDS    = z(RSDS),
    z_soil_pH = z(soil_pH),
    z_Bee     = z(Bee_Richness)
  )

stopifnot(nrow(dat0) >= 50)
cat("✓ dat0 ready. n =", nrow(dat0), "\n")
```

### 8. Check predictor correlations and choose the competing pair automatically

```{r select-competing-pair}
main_effect_pool <- c(
  "topo_PC1",
  "Temperature_PC1",
  "precip_PC1",
  "soil_nutrient_PC1",
  "soil_phys_PC1",
  "z_soil_pH",
  "z_RSDS"
)

replaceable_pool <- setdiff(
  main_effect_pool,
  c("topo_PC1", "Temperature_PC1", "z_RSDS")
)

cor_mat <- cor(dat0[, main_effect_pool], use = "pairwise.complete.obs")
write.csv(cor_mat, file.path(cmp_dir, "predictor_correlation_matrix.csv"), row.names = TRUE)

cor_df <- as.data.frame(as.table(cor_mat), stringsAsFactors = FALSE)
names(cor_df) <- c("var1", "var2", "r")

high_cor <- cor_df %>%
  dplyr::filter(var1 != var2) %>%
  dplyr::mutate(abs_r = abs(r)) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(pair = paste(sort(c(var1, var2)), collapse = "___")) %>%
  dplyr::ungroup() %>%
  dplyr::distinct(pair, .keep_all = TRUE) %>%
  dplyr::filter(var1 %in% replaceable_pool, var2 %in% replaceable_pool) %>%
  dplyr::arrange(desc(abs_r))

write_csv(high_cor, file.path(cmp_dir, "high_correlation_pairs_replaceable_pool.csv"))
print(high_cor)

corr_threshold <- 0.7
candidate_pairs <- high_cor %>%
  dplyr::filter(abs_r >= corr_threshold)

if (nrow(candidate_pairs) == 0) {
  stop("No high-correlation competing pair found in replaceable_pool at threshold = ", corr_threshold)
}

best_pair <- candidate_pairs[1, ]
varA <- best_pair$var1
varB <- best_pair$var2

cat("\n==============================\n")
cat("Selected competing pair automatically:\n")
cat("varA =", varA, "\n")
cat("varB =", varB, "\n")
cat("r    =", round(best_pair$r, 3), "\n")
cat("==============================\n\n")

write_csv(best_pair, file.path(cmp_dir, "selected_competing_pair.csv"))
```

### 9. Add interaction terms

```{r add-interactions}
nm_int_A_topo    <- paste0("int_", varA, "_topo")
nm_int_A_rsds    <- paste0("int_", varA, "_rsds")
nm_int_B_topo    <- paste0("int_", varB, "_topo")
nm_int_B_rsds    <- paste0("int_", varB, "_rsds")
nm_int_topo_rsds <- "int_topo_rsds"

dat_cmp <- dat0
dat_cmp[[nm_int_A_topo]]    <- dat_cmp[[varA]] * dat_cmp[["topo_PC1"]]
dat_cmp[[nm_int_A_rsds]]    <- dat_cmp[[varA]] * dat_cmp[["z_RSDS"]]
dat_cmp[[nm_int_B_topo]]    <- dat_cmp[[varB]] * dat_cmp[["topo_PC1"]]
dat_cmp[[nm_int_B_rsds]]    <- dat_cmp[[varB]] * dat_cmp[["z_RSDS"]]
dat_cmp[[nm_int_topo_rsds]] <- dat_cmp[["topo_PC1"]] * dat_cmp[["z_RSDS"]]
```

### 10. Define candidate model sets

```{r define-model-sets}
bee_var <- "z_Bee"

make_full_terms <- function(chosen_var, include_bee = FALSE) {
  other_terms <- setdiff(main_effect_pool, c(varA, varB))
  terms <- c("topo_PC1", "Temperature_PC1", chosen_var,
             setdiff(other_terms, c("topo_PC1", "Temperature_PC1")))
  terms <- unique(terms)
  if (include_bee) terms <- c(terms, bee_var)
  terms
}

full_A_nobee <- make_full_terms(varA, include_bee = FALSE)
full_A_bee   <- make_full_terms(varA, include_bee = TRUE)
full_B_nobee <- make_full_terms(varB, include_bee = FALSE)
full_B_bee   <- make_full_terms(varB, include_bee = TRUE)

candidate_terms <- list(
  A_full_nobee = full_A_nobee,
  A_full_bee   = full_A_bee,
  B_full_nobee = full_B_nobee,
  B_full_bee   = full_B_bee,
  A_int_topo      = c(full_A_bee, nm_int_A_topo),
  A_int_rsds      = c(full_A_bee, nm_int_A_rsds),
  A_int_topo_rsds = c(full_A_bee, nm_int_topo_rsds),
  B_int_topo      = c(full_B_bee, nm_int_B_topo),
  B_int_rsds      = c(full_B_bee, nm_int_B_rsds),
  B_int_topo_rsds = c(full_B_bee, nm_int_topo_rsds)
)

candidate_family <- c(
  A_full_nobee = "main_effect_models",
  A_full_bee   = "main_effect_models",
  B_full_nobee = "main_effect_models",
  B_full_bee   = "main_effect_models",
  A_int_topo      = "interaction_models",
  A_int_rsds      = "interaction_models",
  A_int_topo_rsds = "interaction_models",
  B_int_topo      = "interaction_models",
  B_int_rsds      = "interaction_models",
  B_int_topo_rsds = "interaction_models"
)

model_definition_tbl <- tibble(
  model = names(candidate_terms),
  family = candidate_family[names(candidate_terms)],
  terms = vapply(candidate_terms, function(x) paste(x, collapse = " + "), character(1))
)
write_csv(model_definition_tbl, file.path(cmp_dir, "model_definitions.csv"))
print(model_definition_tbl)
```
### 11. Build the mesh and SPDE object

```{r mesh-spde}
sf_cmp <- sf::st_as_sf(dat_cmp, coords = c("longitude", "latitude"), crs = 4326)
sf_cmp_m <- sf::st_transform(sf_cmp, 3857)
coords_cmp <- sf::st_coordinates(sf_cmp_m)

mesh_cmp <- INLA::inla.mesh.2d(
  loc      = coords_cmp,
  max.edge = c(20000, 100000),
  cutoff   = 5000
)

spde_cmp <- INLA::inla.spde2.matern(mesh_cmp, alpha = 2)
A_cmp    <- INLA::inla.spde.make.A(mesh_cmp, loc = coords_cmp)

cat("✓ comparison mesh built.\n")
```

### 12. Define model-fitting helper functions

```{r fit-functions}
fit_spde_model <- function(dat, spde, A, terms, model_name, family_name) {
  stopifnot(all(terms %in% names(dat)))

  X <- dat[, terms, drop = FALSE]
  X$Intercept <- 1
  X <- X[, c("Intercept", terms), drop = FALSE]

  stk <- INLA::inla.stack(
    data = list(y = dat$y),
    A = list(A, 1),
    effects = list(
      s = 1:spde$n.spde,
      X = X
    ),
    tag = "est"
  )

  rhs <- paste(c("0 + Intercept", terms, "f(s, model = spde)"), collapse = " + ")
  form <- as.formula(paste("y ~", rhs))

  fit <- INLA::inla(
    formula = form,
    data    = INLA::inla.stack.data(stk),
    family  = "gaussian",
    control.predictor = list(
      A = INLA::inla.stack.A(stk),
      compute = TRUE
    ),
    control.compute = list(
      config = TRUE,
      dic  = TRUE,
      waic = TRUE,
      cpo  = TRUE
    )
  )

  list(
    fit = fit,
    stk = stk,
    formula = form,
    model_name = model_name,
    family_name = family_name,
    terms = terms
  )
}

get_model_metrics <- function(obj) {
  fit <- obj$fit
  cpo_vals <- fit$cpo$cpo

  tibble(
    model = obj$model_name,
    family = obj$family_name,
    n_terms = length(obj$terms),
    terms = paste(obj$terms, collapse = " + "),
    DIC = fit$dic$dic,
    WAIC = fit$waic$waic,
    pWAIC = fit$waic$p.eff,
    mean_neglogCPO = mean(-log(cpo_vals), na.rm = TRUE),
    n_CPO_NA = sum(is.na(cpo_vals))
  )
}
```
### 13. Fit all candidate SPDE-INLA models

```{r fit-all-models}
fit_list <- lapply(names(candidate_terms), function(nm) {
  cat("Fitting:", nm, "\n")
  fit_spde_model(
    dat = dat_cmp,
    spde = spde_cmp,
    A = A_cmp,
    terms = candidate_terms[[nm]],
    model_name = nm,
    family_name = candidate_family[[nm]]
  )
})
names(fit_list) <- names(candidate_terms)
```

### 14. Compare candidate models

```{r compare-models}
cmp_tbl <- bind_rows(lapply(fit_list, get_model_metrics)) %>%
  arrange(WAIC) %>%
  mutate(
    delta_WAIC = WAIC - min(WAIC, na.rm = TRUE),
    delta_DIC  = DIC  - min(DIC,  na.rm = TRUE),
    rank_WAIC  = row_number()
  )

print(cmp_tbl, n = nrow(cmp_tbl))
write_csv(cmp_tbl, file.path(cmp_dir, "auto_competing_pair_model_comparison.csv"))

family_best <- cmp_tbl %>%
  group_by(family) %>%
  slice_min(order_by = WAIC, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(WAIC) %>%
  mutate(delta_WAIC_global = WAIC - min(WAIC))

print(family_best)
write_csv(family_best, file.path(cmp_dir, "auto_competing_pair_family_best.csv"))

top_models <- cmp_tbl %>%
  dplyr::filter(delta_WAIC < 2) %>%
  dplyr::arrange(WAIC, n_terms)

print(top_models)
write_csv(top_models, file.path(cmp_dir, "auto_competing_pair_top_models_deltaWAIC_lt2.csv"))

best_model_name <- cmp_tbl$model[1]
best_obj <- fit_list[[best_model_name]]

cat("\n==============================\n")
cat("Best model by WAIC:", best_model_name, "\n")
cat("Selected pair:", varA, "vs", varB, "\n")
cat("Formula:\n")
print(best_obj$formula)
cat("==============================\n")
```

### 15. Save model comparison plots

```{r comparison-plots}
waic_range <- range(cmp_tbl$WAIC, na.rm = TRUE)
waic_span  <- diff(waic_range)
if (!is.finite(waic_span) || waic_span == 0) waic_span <- 1

p_cmp <- cmp_tbl %>%
  mutate(model = forcats::fct_reorder(model, WAIC)) %>%
  ggplot(aes(x = WAIC, y = model, shape = family)) +
  geom_point(size = 3) +
  geom_text(
    aes(label = paste0("Δ=", round(delta_WAIC, 2))),
    nudge_x = 0.03 * waic_span,
    size = 3
  ) +
  labs(
    x = "WAIC",
    y = NULL,
    title = "Candidate model comparison",
    subtitle = paste0("Selected competing pair: ", varA, " vs ", varB)
  ) +
  theme_bw(base_size = 12)

print(p_cmp)

ggsave(
  filename = file.path(cmp_dir, "candidate_model_comparison_WAIC.png"),
  plot = p_cmp,
  width = 9,
  height = 6.5,
  dpi = 300
)

p_delta <- cmp_tbl %>%
  mutate(model = forcats::fct_reorder(model, delta_WAIC)) %>%
  ggplot(aes(x = delta_WAIC, y = model, fill = family)) +
  geom_col() +
  geom_vline(xintercept = c(2, 7, 10), linetype = 2, color = "grey60") +
  labs(
    x = "ΔWAIC",
    y = NULL,
    title = "Relative model support",
    subtitle = "Rough guide: ΔWAIC < 2 similar; > 10 much worse"
  ) +
  theme_bw(base_size = 12)

print(p_delta)

ggsave(
  filename = file.path(cmp_dir, "candidate_model_comparison_deltaWAIC.png"),
  plot = p_delta,
  width = 9,
  height = 6.5,
  dpi = 300
)
```

### 16. Save formulas and fixed-effect tables

```{r save-formulas-fixef}
formula_tbl <- tibble(
  model = names(fit_list),
  family = vapply(fit_list, function(x) x$family_name, character(1)),
  formula = vapply(fit_list, function(x) paste(deparse(x$formula), collapse = " "), character(1))
)
write_csv(formula_tbl, file.path(cmp_dir, "auto_competing_pair_model_formulas.csv"))

fixef_all <- bind_rows(lapply(fit_list, function(obj) {
  obj$fit$summary.fixed %>%
    as.data.frame() %>%
    tibble::rownames_to_column("term") %>%
    mutate(model = obj$model_name, family = obj$family_name, .before = 1)
}))
write_csv(fixef_all, file.path(cmp_dir, "auto_competing_pair_all_fixed_effects.csv"))
```

### 17. Refit all top-supported models for supplementary outputs

```{r refit-top-models}
refit_one_model <- function(obj, dat, coords, mesh, spde, A, outdir) {
  terms <- obj$terms

  X <- dat[, terms, drop = FALSE]
  X$Intercept <- 1
  X <- X[, c("Intercept", terms), drop = FALSE]

  stk <- INLA::inla.stack(
    data = list(y = dat$y),
    A = list(A, 1),
    effects = list(
      s = 1:spde$n.spde,
      X = X
    ),
    tag = "est"
  )

  fit <- INLA::inla(
    formula = obj$formula,
    data    = INLA::inla.stack.data(stk),
    family  = "gaussian",
    control.predictor = list(
      A = INLA::inla.stack.A(stk),
      compute = TRUE
    ),
    control.compute = list(
      config = TRUE,
      dic  = TRUE,
      waic = TRUE,
      cpo  = TRUE
    )
  )

  model_dir <- file.path(outdir, obj$model_name)
  dir.create(model_dir, recursive = TRUE, showWarnings = FALSE)

  fixef <- fit$summary.fixed %>%
    as.data.frame() %>%
    tibble::rownames_to_column("term")

  write_csv(fixef, file.path(model_dir, "fixed_effects.csv"))

  fit_stats <- tibble(
    model = obj$model_name,
    family = obj$family_name,
    DIC = fit$dic$dic,
    WAIC = fit$waic$waic,
    pWAIC = fit$waic$p.eff,
    mean_neglogCPO = mean(-log(fit$cpo$cpo), na.rm = TRUE),
    n_CPO_NA = sum(is.na(fit$cpo$cpo))
  )
  write_csv(fit_stats, file.path(model_dir, "fit_stats.csv"))

  IDX <- INLA::inla.stack.index(stk, "est")$data
  y_obs   <- INLA::inla.stack.data(stk)$y[IDX]
  yhat    <- fit$summary.fitted.values$mean[IDX]
  yhat_sd <- fit$summary.fitted.values$sd[IDX]

  resid_raw  <- y_obs - yhat
  resid_zfit <- resid_raw / pmax(as.numeric(yhat_sd), 1e-6)

  dat_res <- dat
  dat_res$yhat       <- yhat
  dat_res$yhat_sd    <- yhat_sd
  dat_res$resid_raw  <- resid_raw
  dat_res$resid_zfit <- resid_zfit

  write_csv(dat_res, file.path(model_dir, "residuals.csv"))

  coef_plot_tbl <- fixef %>%
    dplyr::filter(term != "Intercept") %>%
    dplyr::mutate(
      signif95 = (`0.025quant` > 0) | (`0.975quant` < 0),
      sig_label = ifelse(signif95, "95% CI excludes 0", "95% CI overlaps 0"),
      term_pretty = forcats::fct_reorder(term, mean)
    )

  p_fixef <- ggplot(coef_plot_tbl,
                    aes(x = mean, y = term_pretty, shape = sig_label)) +
    geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
    geom_errorbarh(aes(xmin = `0.025quant`, xmax = `0.975quant`), height = 0.18) +
    geom_point(size = 3) +
    labs(
      x = "Posterior mean coefficient",
      y = NULL,
      shape = NULL,
      title = paste0("Fixed effects: ", obj$model_name),
      subtitle = paste0("family = ", obj$family_name)
    ) +
    theme_bw(base_size = 12)

  print(p_fixef)

  ggsave(
    filename = file.path(model_dir, "fixed_effects_plot.png"),
    plot = p_fixef,
    width = 8,
    height = 5.5,
    dpi = 300
  )

  write_csv(coef_plot_tbl, file.path(model_dir, "fixed_effects_for_plot.csv"))

  invisible(list(
    fit = fit,
    stk = stk,
    fixef = fixef,
    fit_stats = fit_stats,
    residuals = dat_res,
    coef_plot_tbl = coef_plot_tbl
  ))
}

top_model_names <- top_models$model
top_model_objects <- fit_list[top_model_names]

top_refits <- lapply(top_model_objects, refit_one_model,
                     dat = dat_cmp,
                     coords = coords_cmp,
                     mesh = mesh_cmp,
                     spde = spde_cmp,
                     A = A_cmp,
                     outdir = top_dir)

names(top_refits) <- top_model_names

write_csv(
  tibble(model = top_model_names),
  file.path(top_dir, "top_model_names.csv")
)
```

### 18. Plot only the top-supported models

```{r top-model-plots}
top_cmp_tbl <- cmp_tbl %>%
  dplyr::filter(model %in% top_model_names) %>%
  dplyr::arrange(WAIC)

top_waic_range <- range(top_cmp_tbl$WAIC, na.rm = TRUE)
top_waic_span  <- diff(top_waic_range)
if (!is.finite(top_waic_span) || top_waic_span == 0) top_waic_span <- 1

p_top <- top_cmp_tbl %>%
  mutate(model = forcats::fct_reorder(model, WAIC)) %>%
  ggplot(aes(x = WAIC, y = model, shape = family)) +
  geom_point(size = 3) +
  geom_text(
    aes(label = paste0("Δ=", round(delta_WAIC, 2))),
    nudge_x = 0.03 * top_waic_span,
    size = 3
  ) +
  labs(
    x = "WAIC",
    y = NULL,
    title = "Top-supported models",
    subtitle = "Models with ΔWAIC < 2"
  ) +
  theme_bw(base_size = 12)

print(p_top)

ggsave(
  filename = file.path(top_dir, "top_models_WAIC.png"),
  plot = p_top,
  width = 8,
  height = 4.8,
  dpi = 300
)
```

### 19. Refit the best model cleanly

```{r refit-best-model}
best_terms <- best_obj$terms

dat_final <- dat_cmp
coords_final <- coords_cmp
mesh_final <- mesh_cmp
spde_final <- spde_cmp
A_final <- A_cmp

X_final <- dat_final[, best_terms, drop = FALSE]
X_final$Intercept <- 1
X_final <- X_final[, c("Intercept", best_terms), drop = FALSE]

stk_final <- INLA::inla.stack(
  data = list(y = dat_final$y),
  A = list(A_final, 1),
  effects = list(
    s = 1:spde_final$n.spde,
    X = X_final
  ),
  tag = "est"
)

formula_final <- best_obj$formula

fit_final <- INLA::inla(
  formula = formula_final,
  data    = INLA::inla.stack.data(stk_final),
  family  = "gaussian",
  control.predictor = list(
    A = INLA::inla.stack.A(stk_final),
    compute = TRUE
  ),
  control.compute = list(
    config = TRUE,
    dic  = TRUE,
    waic = TRUE,
    cpo  = TRUE
  )
)

cat("✓ best model refitted.\n")
```

### 20. Save fixed effects, variance decomposition, and residuals for the best model

```{r best-model-results}
fixef_final <- fit_final$summary.fixed %>%
  as.data.frame() %>%
  tibble::rownames_to_column("term")

write_csv(fixef_final, file.path(best_dir, "best_model_fixed_effects.csv"))
print(fixef_final)

IDX <- INLA::inla.stack.index(stk_final, "est")$data

proj_obs <- INLA::inla.mesh.projector(mesh_final, loc = coords_final)
s_obs <- as.numeric(INLA::inla.mesh.project(proj_obs, fit_final$summary.random$s$mean))

eta <- fit_final$summary.linear.predictor$mean[IDX]
fixed_part <- eta - s_obs

prec_med <- fit_final$summary.hyperpar["Precision for the Gaussian observations", "0.5quant"]
resid_var <- 1 / prec_med

v_fixed   <- var(fixed_part, na.rm = TRUE)
v_spatial <- var(s_obs, na.rm = TRUE)
v_resid   <- resid_var
v_total   <- v_fixed + v_spatial + v_resid

var_decomp <- data.frame(
  component  = c("fixed", "spatial", "residual"),
  variance   = c(v_fixed, v_spatial, v_resid),
  proportion = c(v_fixed, v_spatial, v_resid) / v_total
)
write_csv(var_decomp, file.path(best_dir, "best_model_variance_decomposition.csv"))
print(var_decomp)

sh <- fit_final$summary.hyperpar
tau   <- exp(sh[grep("Theta1", rownames(sh)), "0.5quant"])
kappa <- exp(sh[grep("Theta2", rownames(sh)), "0.5quant"])
range_km <- sqrt(8) / kappa / 1000
sigma    <- 1 / (sqrt(4*pi) * tau * kappa)

spatial_pars <- data.frame(range_km = range_km, sigma = sigma)
write_csv(spatial_pars, file.path(best_dir, "best_model_spatial_parameters.csv"))
print(spatial_pars)

y_obs    <- INLA::inla.stack.data(stk_final)$y[IDX]
yhat     <- fit_final$summary.fitted.values$mean[IDX]
yhat_sd  <- fit_final$summary.fitted.values$sd[IDX]

resid_raw  <- y_obs - yhat
resid_zfit <- resid_raw / pmax(as.numeric(yhat_sd), 1e-6)

dat_used <- dat_final
dat_used$yhat       <- yhat
dat_used$yhat_sd    <- yhat_sd
dat_used$resid_raw  <- resid_raw
dat_used$resid_zfit <- resid_zfit

write_csv(dat_used, file.path(best_dir, "best_model_residuals.csv"))
```

### 21. Plot fixed effects for the best model

```{r best-model-fixef-plot}
coef_plot_tbl <- fixef_final %>%
  dplyr::filter(term != "Intercept") %>%
  dplyr::mutate(
    signif95 = (`0.025quant` > 0) | (`0.975quant` < 0),
    sig_label = ifelse(signif95, "95% CI excludes 0", "95% CI overlaps 0")
  )

p_fixef <- ggplot(coef_plot_tbl,
                  aes(x = mean, y = forcats::fct_reorder(term, mean), shape = sig_label)) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
  geom_errorbarh(aes(xmin = `0.025quant`, xmax = `0.975quant`), height = 0.18) +
  geom_point(size = 3) +
  labs(
    x = "Posterior mean coefficient",
    y = NULL,
    shape = NULL,
    title = "Fixed effects in the best auto-selected SPDE-INLA model",
    subtitle = paste0("Best model: ", best_model_name, " | pair: ", varA, " vs ", varB)
  ) +
  theme_bw(base_size = 12)

print(p_fixef)

ggsave(
  filename = file.path(best_plotdir, "best_model_fixed_effects.png"),
  plot = p_fixef,
  width = 8.2,
  height = 5.8,
  dpi = 300
)

write_csv(coef_plot_tbl, file.path(best_dir, "best_model_fixed_effects_for_plot.csv"))
```

### 22. Save best-model formula and fit statistics

```{r best-model-formula-fitstats}
formula_best_tbl <- tibble(
  selected_varA = varA,
  selected_varB = varB,
  model = best_model_name,
  family = best_obj$family_name,
  formula = paste(deparse(formula_final), collapse = " ")
)

fit_stats <- tibble(
  selected_varA = varA,
  selected_varB = varB,
  model = best_model_name,
  family = best_obj$family_name,
  DIC = fit_final$dic$dic,
  WAIC = fit_final$waic$waic,
  pWAIC = fit_final$waic$p.eff,
  mean_neglogCPO = mean(-log(fit_final$cpo$cpo), na.rm = TRUE),
  n_CPO_NA = sum(is.na(fit_final$cpo$cpo))
)

write_csv(formula_best_tbl, file.path(best_dir, "best_model_formula.csv"))
write_csv(fit_stats,        file.path(best_dir, "best_model_fit_stats.csv"))
print(fit_stats)
```

### 23. Fit qGAM models using the residuals from the best model

```{r qgam-main}
set.seed(42)

MODE <- "fit"
POSITIVE_THRESH <- 1.5

K_SPACE <- 60
K_ELEV  <- 10
K_PD    <- 10
K_DOY   <- 12
K_INT2  <- c(8, 8)

ADD_DOYxPD <- FALSE
taus <- c(0.75, 0.85, 0.90, 0.95)

q_lo <- 0.10
q_hi <- 0.90
SPACE_AVERAGE <- "marginal"
FIXED_LONLAT  <- c(longitude = NA_real_, latitude = NA_real_)

if ("date" %in% names(dat_used)) {
  date_parsed <- suppressWarnings(as.Date(
    dat_used$date,
    tryFormats = c(
      "%Y-%m-%d","%Y/%m/%d","%Y.%m.%d","%Y%m%d",
      "%m/%d/%Y","%m-%d-%Y",
      "%Y-%m-%d %H:%M","%Y/%m/%d %H:%M",
      "%Y-%m-%d %H:%M:%S","%Y/%m/%d %H:%M:%S"
    )
  ))
  dat_used$DOY <- as.integer(strftime(date_parsed, "%j"))
} else {
  dat_used$DOY <- NA_integer_
}

human_tifs <- list.files(human_dir, pattern = "\\.tif$", full.names = TRUE)
stopifnot(length(human_tifs) > 0)

r_human <- terra::rast(human_tifs[1])

pts_ll <- terra::vect(
  data.frame(lon = dat_used$longitude, lat = dat_used$latitude),
  geom = c("lon", "lat"),
  crs  = "EPSG:4326"
)

if (!terra::same.crs(pts_ll, r_human)) {
  pts_ll <- terra::project(pts_ll, terra::crs(r_human))
}

vals <- terra::extract(r_human, pts_ll, ID = FALSE)[, 1]
dat_used$pop_density_raw <- vals
dat_used$pop_density_log <- log1p(vals)

if (!("elevation" %in% names(dat_used))) {
  dat_used <- dat_used %>%
    left_join(
      df2 %>%
        dplyr::select(longitude, latitude, elevation) %>%
        distinct(),
      by = c("longitude", "latitude")
    )
}

stopifnot("elevation" %in% names(dat_used))
elev0 <- median(dat_used$elevation, na.rm = TRUE)

dat_used <- dat_used %>%
  dplyr::filter(
    is.finite(resid_zfit),
    is.finite(pop_density_log),
    is.finite(longitude),
    is.finite(latitude),
    is.finite(elevation)
  )

df_fit  <- dat_used
df_dark <- dat_used %>% dplyr::filter(resid_zfit > POSITIVE_THRESH)

df_fit_used <- if (MODE == "dark") df_dark else df_fit
df_pred     <- df_fit_used

use_DOY <- all(is.finite(df_fit_used$DOY)) && length(unique(df_fit_used$DOY)) >= 8

write_csv(df_fit_used, file.path(gam_dir, paste0("df_used_MODE_", MODE, ".csv")))

base_terms <- paste0(
  "s(longitude, latitude, k=", K_SPACE, ") + ",
  "s(elevation, k=", K_ELEV, ") + ",
  "s(pop_density_log, k=", K_PD, ")"
)

doy_terms <- if (use_DOY) {
  if (ADD_DOYxPD) {
    paste0(
      " + s(DOY, bs='tp', k=", K_DOY, ")",
      " + ti(DOY, pop_density_log, k=c(", K_INT2[1], ",", K_INT2[2], "))"
    )
  } else {
    paste0(" + s(DOY, bs='tp', k=", K_DOY, ")")
  }
} else ""

form_q <- as.formula(paste0("resid_zfit ~ ", base_terms, doy_terms))
print(form_q)

mods <- lapply(taus, function(tau) qgam::qgam(form_q, data = df_fit_used, qu = tau))
names(mods) <- paste0("q", taus)

sink(file.path(gam_dir, paste0("qGAM_summary_MODE_", MODE, ".txt")))
cat("Selected competing pair =", varA, "vs", varB, "\n")
cat("Best SPDE model =", best_model_name, "\n")
cat("Best SPDE family =", best_obj$family_name, "\n\n")
cat("===== FORMULA =====\n")
print(form_q)
cat("\n")
for (nm in names(mods)) {
  cat("-----", nm, "-----\n")
  print(summary(mods[[nm]]))
  cat("\n")
}
sink()

smooth_tbl <- bind_rows(lapply(names(mods), function(nm) {
  sm <- summary(mods[[nm]])
  st <- as.data.frame(sm$s.table)
  st$term <- rownames(st)
  rownames(st) <- NULL
  names(st)[1:4] <- c("edf", "Ref.df", "Chi.sq", "p_value")
  tibble::as_tibble(st) %>%
    mutate(
      tau = nm,
      formula = paste(deparse(form_q), collapse = " ")
    )
}))
write_csv(smooth_tbl, file.path(gam_dir, paste0("qGAM_smooth_table_MODE_", MODE, ".csv")))
```

### 24. Define qGAM effect-size helper functions

```{r qgam-helpers}
make_newdata_pair <- function(df_space,
                              elev0, DOY0,
                              pd_lo, pd_hi,
                              doy_lo, doy_hi,
                              space_average = c("marginal", "fixed"),
                              fixed_lonlat = c(longitude = NA_real_, latitude = NA_real_)) {
  space_average <- match.arg(space_average)

  nd_pd_lo <- df_space
  nd_pd_hi <- df_space
  nd_pd_lo$elevation <- elev0
  nd_pd_hi$elevation <- elev0
  nd_pd_lo$pop_density_log <- pd_lo
  nd_pd_hi$pop_density_log <- pd_hi
  if ("DOY" %in% names(df_space)) {
    nd_pd_lo$DOY <- DOY0
    nd_pd_hi$DOY <- DOY0
  }

  pd_mid <- mean(c(pd_lo, pd_hi))
  nd_doy_lo <- df_space
  nd_doy_hi <- df_space
  nd_doy_lo$elevation <- elev0
  nd_doy_hi$elevation <- elev0
  nd_doy_lo$pop_density_log <- pd_mid
  nd_doy_hi$pop_density_log <- pd_mid
  if ("DOY" %in% names(df_space)) {
    nd_doy_lo$DOY <- doy_lo
    nd_doy_hi$DOY <- doy_hi
  }

  if (space_average == "fixed") {
    lon0 <- fixed_lonlat[["longitude"]]
    lat0 <- fixed_lonlat[["latitude"]]
    if (!is.finite(lon0)) lon0 <- median(df_space$longitude, na.rm = TRUE)
    if (!is.finite(lat0)) lat0 <- median(df_space$latitude,  na.rm = TRUE)

    nd_pd_lo$longitude  <- lon0
    nd_pd_lo$latitude   <- lat0
    nd_pd_hi$longitude  <- lon0
    nd_pd_hi$latitude   <- lat0
    nd_doy_lo$longitude <- lon0
    nd_doy_lo$latitude  <- lat0
    nd_doy_hi$longitude <- lon0
    nd_doy_hi$latitude  <- lat0
  }

  list(
    nd_pd_lo  = nd_pd_lo,
    nd_pd_hi  = nd_pd_hi,
    nd_doy_lo = nd_doy_lo,
    nd_doy_hi = nd_doy_hi
  )
}

deltaQ_PD <- function(mod, df_space, elev0, DOY0, pd_lo, pd_hi,
                      space_average = SPACE_AVERAGE,
                      fixed_lonlat = FIXED_LONLAT) {
  nd <- make_newdata_pair(
    df_space,
    elev0 = elev0, DOY0 = DOY0,
    pd_lo = pd_lo, pd_hi = pd_hi,
    doy_lo = NA_real_, doy_hi = NA_real_,
    space_average = space_average,
    fixed_lonlat = fixed_lonlat
  )
  q_lo_hat <- mean(predict(mod, newdata = nd$nd_pd_lo, type = "response"), na.rm = TRUE)
  q_hi_hat <- mean(predict(mod, newdata = nd$nd_pd_hi, type = "response"), na.rm = TRUE)
  c(deltaQ_PD = q_hi_hat - q_lo_hat, Q_lo = q_lo_hat, Q_hi = q_hi_hat)
}

deltaQ_DOY <- function(mod, df_space, elev0, pd0, doy_lo, doy_hi,
                       space_average = SPACE_AVERAGE,
                       fixed_lonlat = FIXED_LONLAT) {
  nd_lo <- df_space
  nd_hi <- df_space
  nd_lo$elevation <- elev0
  nd_hi$elevation <- elev0
  nd_lo$pop_density_log <- pd0
  nd_hi$pop_density_log <- pd0
  nd_lo$DOY <- doy_lo
  nd_hi$DOY <- doy_hi

  if (space_average == "fixed") {
    lon0 <- fixed_lonlat[["longitude"]]
    lat0 <- fixed_lonlat[["latitude"]]
    if (!is.finite(lon0)) lon0 <- median(df_space$longitude, na.rm = TRUE)
    if (!is.finite(lat0)) lat0 <- median(df_space$latitude,  na.rm = TRUE)
    nd_lo$longitude <- lon0
    nd_lo$latitude  <- lat0
    nd_hi$longitude <- lon0
    nd_hi$latitude  <- lat0
  }

  q_lo_hat <- mean(predict(mod, newdata = nd_lo, type = "response"), na.rm = TRUE)
  q_hi_hat <- mean(predict(mod, newdata = nd_hi, type = "response"), na.rm = TRUE)
  c(deltaQ_DOY = q_hi_hat - q_lo_hat, Q_lo = q_lo_hat, Q_hi = q_hi_hat)
}

pd_lo0  <- as.numeric(quantile(df_fit_used$pop_density_log, q_lo, na.rm = TRUE))
pd_hi0  <- as.numeric(quantile(df_fit_used$pop_density_log, q_hi, na.rm = TRUE))
pd_mid0 <- mean(c(pd_lo0, pd_hi0))

DOY0    <- if (use_DOY) median(df_fit_used$DOY, na.rm = TRUE) else NA_real_
doy_lo0 <- if (use_DOY) as.numeric(quantile(df_fit_used$DOY, q_lo, na.rm = TRUE)) else NA_real_
doy_hi0 <- if (use_DOY) as.numeric(quantile(df_fit_used$DOY, q_hi, na.rm = TRUE)) else NA_real_

point_PD <- lapply(taus, function(tau) {
  mod <- mods[[paste0("q", tau)]]
  out <- deltaQ_PD(
    mod,
    df_space = df_pred,
    elev0 = elev0,
    DOY0 = DOY0,
    pd_lo = pd_lo0,
    pd_hi = pd_hi0
  )
  data.frame(
    tau = tau,
    deltaQ_PD = as.numeric(out["deltaQ_PD"]),
    Q_lo_PD   = as.numeric(out["Q_lo"]),
    Q_hi_PD   = as.numeric(out["Q_hi"])
  )
}) %>% bind_rows()

point_DOY <- if (use_DOY) {
  lapply(taus, function(tau) {
    mod <- mods[[paste0("q", tau)]]
    out <- deltaQ_DOY(
      mod,
      df_space = df_pred,
      elev0 = elev0,
      pd0 = pd_mid0,
      doy_lo = doy_lo0,
      doy_hi = doy_hi0
    )
    data.frame(
      tau = tau,
      deltaQ_DOY = as.numeric(out["deltaQ_DOY"]),
      Q_lo_DOY   = as.numeric(out["Q_lo"]),
      Q_hi_DOY   = as.numeric(out["Q_hi"])
    )
  }) %>% bind_rows()
} else {
  data.frame(
    tau = taus,
    deltaQ_DOY = NA_real_,
    Q_lo_DOY = NA_real_,
    Q_hi_DOY = NA_real_
  )
}

point_DOY_by_PD <- if (use_DOY) {
  lapply(taus, function(tau) {
    mod <- mods[[paste0("q", tau)]]
    lo <- deltaQ_DOY(
      mod,
      df_space = df_pred,
      elev0 = elev0,
      pd0 = pd_lo0,
      doy_lo = doy_lo0,
      doy_hi = doy_hi0
    )
    hi <- deltaQ_DOY(
      mod,
      df_space = df_pred,
      elev0 = elev0,
      pd0 = pd_hi0,
      doy_lo = doy_lo0,
      doy_hi = doy_hi0
    )
    data.frame(
      tau = tau,
      deltaQ_DOY_lowPD  = as.numeric(lo["deltaQ_DOY"]),
      deltaQ_DOY_highPD = as.numeric(hi["deltaQ_DOY"]),
      diff              = as.numeric(hi["deltaQ_DOY"]) - as.numeric(lo["deltaQ_DOY"])
    )
  }) %>% bind_rows()
} else {
  data.frame(
    tau = taus,
    deltaQ_DOY_lowPD = NA_real_,
    deltaQ_DOY_highPD = NA_real_,
    diff = NA_real_
  )
}

res_point <- point_PD %>%
  left_join(point_DOY, by = "tau") %>%
  left_join(point_DOY_by_PD, by = "tau") %>%
  mutate(
    pd_lo = pd_lo0,
    pd_hi = pd_hi0,
    DOY0 = DOY0,
    elev0 = elev0,
    space_average = SPACE_AVERAGE,
    MODE = MODE
  )

write_csv(res_point, file.path(gam_dir, paste0("point_effect_sizes_MODE_", MODE, ".csv")))
print(res_point)
```

### 25. Plot qGAM smooths

```{r qgam-effect-plots}
plot_smooth_1d <- function(mod, data, varname, file_stub, n = 200) {
  xseq <- seq(
    min(data[[varname]], na.rm = TRUE),
    max(data[[varname]], na.rm = TRUE),
    length.out = n
  )

  nd <- data[rep(1, n), , drop = FALSE]
  nd[[varname]] <- xseq

  if ("longitude" %in% names(nd)) nd$longitude <- median(data$longitude, na.rm = TRUE)
  if ("latitude"  %in% names(nd)) nd$latitude  <- median(data$latitude,  na.rm = TRUE)
  if ("elevation" %in% names(nd) && varname != "elevation") nd$elevation <- median(data$elevation, na.rm = TRUE)
  if ("pop_density_log" %in% names(nd) && varname != "pop_density_log") nd$pop_density_log <- median(data$pop_density_log, na.rm = TRUE)
  if ("DOY" %in% names(nd) && varname != "DOY" && all(is.finite(data$DOY))) nd$DOY <- median(data$DOY, na.rm = TRUE)

  pr <- predict(mod, newdata = nd, se.fit = TRUE, type = "response")
  plot_df <- data.frame(
    x = xseq,
    fit = as.numeric(pr$fit),
    se  = as.numeric(pr$se.fit)
  ) %>%
    mutate(
      lo = fit - 1.96 * se,
      hi = fit + 1.96 * se
    )

  p <- ggplot(plot_df, aes(x = x, y = fit)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2) +
    geom_line() +
    theme_bw(base_size = 12) +
    labs(x = varname, y = "Predicted upper-tail residual")

  print(p)

  ggsave(
    filename = file.path(gam_dir, paste0(file_stub, ".png")),
    plot = p,
    width = 5.8,
    height = 4.2,
    dpi = 300
  )

  write_csv(plot_df, file.path(gam_dir, paste0(file_stub, ".csv")))
}

mod_plot <- mods[["q0.95"]]

plot_smooth_1d(mod_plot, df_fit_used, "pop_density_log", "qGAM_q095_pop_density_log")
if (use_DOY) plot_smooth_1d(mod_plot, df_fit_used, "DOY", "qGAM_q095_DOY")
plot_smooth_1d(mod_plot, df_fit_used, "elevation", "qGAM_q095_elevation")
```

### 26. Plot multi-tau qGAM curves

```{r qgam-multi-tau}
get_curve_df <- function(mod, data, varname, n = 200) {
  xseq <- seq(
    min(data[[varname]], na.rm = TRUE),
    max(data[[varname]], na.rm = TRUE),
    length.out = n
  )

  nd <- data[rep(1, n), , drop = FALSE]
  nd[[varname]] <- xseq

  if ("longitude" %in% names(nd)) nd$longitude <- median(data$longitude, na.rm = TRUE)
  if ("latitude"  %in% names(nd)) nd$latitude  <- median(data$latitude,  na.rm = TRUE)
  if ("elevation" %in% names(nd) && varname != "elevation") nd$elevation <- median(data$elevation, na.rm = TRUE)
  if ("pop_density_log" %in% names(nd) && varname != "pop_density_log") nd$pop_density_log <- median(data$pop_density_log, na.rm = TRUE)
  if ("DOY" %in% names(nd) && varname != "DOY" && all(is.finite(data$DOY))) nd$DOY <- median(data$DOY, na.rm = TRUE)

  pr <- predict(mod, newdata = nd, se.fit = TRUE, type = "response")

  data.frame(
    x = xseq,
    fit = as.numeric(pr$fit),
    se  = as.numeric(pr$se.fit)
  ) %>%
    mutate(
      lo = fit - 1.96 * se,
      hi = fit + 1.96 * se
    )
}

plot_multi_tau <- function(mods, data, varname, file_stub) {
  curve_df <- bind_rows(lapply(names(mods), function(nm) {
    get_curve_df(mods[[nm]], data, varname) %>%
      mutate(tau = nm)
  }))

  p <- ggplot(curve_df, aes(x = x, y = fit, colour = tau, fill = tau)) +
    geom_line(linewidth = 1) +
    theme_bw(base_size = 12) +
    labs(x = varname, y = "Predicted upper-tail residual", colour = "tau", fill = "tau")

  print(p)

  ggsave(
    filename = file.path(gam_dir, paste0(file_stub, ".png")),
    plot = p,
    width = 6.4,
    height = 4.6,
    dpi = 300
  )

  write_csv(curve_df, file.path(gam_dir, paste0(file_stub, ".csv")))
}

plot_multi_tau(mods, df_fit_used, "pop_density_log", "qGAM_multitau_pop_density_log")
if (use_DOY) plot_multi_tau(mods, df_fit_used, "DOY", "qGAM_multitau_DOY")
plot_multi_tau(mods, df_fit_used, "elevation", "qGAM_multitau_elevation")
```

### 27. Completion message

```{r completion}
cat("\n✅ COMPLETE PIPELINE DONE\n")
cat("Selected pair:", varA, "vs", varB, "\n")
cat("Comparison dir:", cmp_dir, "\n")
cat("Top-model supp dir:", top_dir, "\n")
cat("Best-model dir:", best_dir, "\n")
cat("qGAM dir:", gam_dir, "\n")
```
